---
title: "Autoantidoby manuscript (Baseline analyses)"
output: html_document
date: "2024-09-20"
editor_options: 
  chunk_output_type: console
---

- in this script I will apply limma model to determine which auto-abs are enrichned in Melanoma vs HD samples
at baseline.
- I will perform differential gene expression
- visualizations: volcano plots, counts 

# Paths/libraries/source code
```{r}
# ---- Project setup: paths, packages, colours/themes ----------------------
# Replaces the original machine-specific absolute paths and the two lost
# source files (autoantibody_libraries.R, LHB_color_tools.R). setup.R defines
# file_path, data_path, results_path and loads all packages + figure themes.
if (file.exists("setup.R")) source("setup.R") else source(file.path("R", "setup.R"))
```

# read PHIPseq_list
```{r}
PHIPseq_list <-  readRDS(paste0(data_path, "PHIPseq_list.Rda"))
names(PHIPseq_list)
```

## select subjects
## define matrix of interest
## & filter background genes
```{r}
VAC_CYC = "_0|_43|_89"
SUBJ_TYPE = c("Melanoma")
```
### prep matrix_phip & meta_phip
```{r}
# Define the function with an additional variable and value for filtering
filter_meta_subjects <- function(meta_name, vac_cycle = NULL, cancer = NULL, outcome = NULL, outcome_val = NULL) {
  # Apply filtering based on input arguments
  
  filtered_df <- PHIPseq_list[[meta_name]] %>%
    filter(
      (is.null(vac_cycle) | grepl(vac_cycle, Vac_cycle)) &
      (is.null(Cancer) | Cancer %in% c(cancer))
    )
  
  # Apply additional filtering if specified by outcome
  if (!is.null(outcome) & !is.null(outcome_val)) {
    filtered_df <- filtered_df %>%
      filter(!!sym(outcome) %in% c(outcome_val))
  }
  
  # Extract and return the 'strings' column
  return(filtered_df %>% pull(strings))
}

# SELECT subjects for analyses
filtered_subjects <- filter_meta_subjects(meta_name = "df_Phip_metadata",
                                           vac_cycle = c(VAC_CYC), 
                                            cancer = c(SUBJ_TYPE)
                                            #outcome = "Outcome", outcome_val = c("Good", "Bad")
                                          )

# Define a function to subset and format matrix and metadata
prepare_phip_data <- function(data_name, phip_list, pts) {
  # Subset the matrix based on the input dataset name and patients of interest
  matrix_phip <- phip_list[[data_name]] %>% 
    dplyr::select(all_of(intersect(pts, colnames(phip_list[[data_name]]))))
  
  # Subset the metadata to match the columns of the matrix
  meta_phip <- phip_list$df_Phip_metadata %>% 
    filter(strings %in% colnames(matrix_phip))
  
  # Set the row names of the metadata to match the column names of the matrix
  rownames(meta_phip) <- meta_phip$strings
  
  # Return the matrix and metadata as a list
  list(matrix = matrix_phip, metadata = meta_phip)
}

# use of the function and CHOOSE dtaaset
phip_data <- prepare_phip_data("Phip_seq_R3_rpk", PHIPseq_list, filtered_subjects)

# Accessing the resulting matrix and metadata
matrix_phip <- phip_data$matrix
matrix_phip <- matrix_phip[rownames(matrix_phip) %in% PHIPseq_list$clear_0count_genes, ] # remove background genes

meta_phip <- phip_data$metadata
####################
if (all(colnames(matrix_phip) == rownames(meta_phip))) {
  print("All matches")
} else {
  print("The column names do not match meta_phip rows !!!")
}

meta_phip %>%  distinct(Subject_ID,Vac_cycle,Outcome)
```

# --------------
# limma model 
```{r}
####################
meta_phip$Outcome <- factor(meta_phip$Outcome, levels=c("Good","Bad"))
meta_phip$Cancer <- factor(meta_phip$Cancer, levels=c("Melanoma","HD"))
####################

library(limma)
# Melanoma vs HD design  
design <- model.matrix(~0 + meta_phip$Cancer, data = meta_phip)
colnames(design) <- c("Melanoma", "HD")

contrasts <- makeContrasts(
  levels = colnames(design),
  Melanoma_vs_HD_baseline = Melanoma - HD
)

log_rpk <- log2(matrix_phip + 0.01)  # Adding 0.01 to avoid log(0)
#log_rpk_normalized <- normalizeBetweenArrays(log_rpk, method = "quantile") if needed
vfit <- lmFit(log_rpk, design)
vfit <- contrasts.fit(vfit, contrasts)
efit <- eBayes(vfit)
toptab <- topTable(efit, sort.by = "P", n = Inf)
toptab$Gene <- rownames(toptab)

limma_toptab <- toptab %>% arrange(logFC) %>%  filter(P.Value <= 0.05 & abs(logFC) > 1)
limma_toptab %<>% mutate(Cancer = if_else(logFC > 0, "Cancer", "Healthy")) 
# ----------------
# CHEKCS
limma_toptab %>% mutate(
  Background = if_else(Gene %in% clear_0count_genes, "Clear", NA)
) %>% janitor::tabyl(Background)
# to check if any of the vaccine antigen peptides may be present
limma_toptab  %>% filter(grepl("TYR|MAGE|MLAN", Gene)) 
```

```{r}
genes_R3 <- limma_toptab
genes_R2 <- limma_toptab

genes_R2_R3 <- intersect(genes_R3$Gene,genes_R2$Gene)
genes_R2_R3 <- genes_R2_R3[!grepl("^LOC", genes_R2_R3)]
# genes overlapping between R3 and R2 lysate enrichment
genes_R2_R3 %>% saveRDS("genes_R2_R3.Rda")
```

### Figures 1 (volcano)
- https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
```{r}
# Gene expression table
fin.volcano.tab <- data.frame(limma_toptab) %>% arrange(desc(logFC))
rownames(fin.volcano.tab) <- fin.volcano.tab$Gene
# up in Melanoma
top_sig_genes_up <- fin.volcano.tab %>% filter(logFC > 0 ) %>% 
  #filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)
# up in HD
top_sig_genes_down <- fin.volcano.tab %>% filter(logFC < 0 ) %>% 
  # filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)

# Volcano Plot
###########
EnhancedVolcano::EnhancedVolcano(toptab,
    lab = rownames(toptab),
    title = "Baseline comparison",
    subtitle = NULL,
    caption = NULL,
    x = 'logFC', maxoverlapsConnectors = Inf,
    y = 'P.Value',
    selectLab = c(top_sig_genes_up,top_sig_genes_down),
    xlab = bquote(~Log[2]~ 'fold change'),
    ylim = c(0, 8),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 3.0,
    labSize = 4.0,
    labCol = 'black',
    labFace = 'plain',
    boxedLabels = TRUE,
    parseLabels = TRUE,
    col = c('black','grey', 'purple', 'red3'),  # Adjust colors if necessary
    colAlpha = 4/5,
    legendLabels = c('Not significant','LogFC > 0.5', 'p-value <= 0.05', 'p-value <= 0.05 & LogFC > 0.5'),
    legendPosition = 'right',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'black') +
    annotate("text", x = c(-3.5, 4.3), y = 8, label = c("Healthy", "Cancer"), size = 8)
```

### Figures 1 (counts)
```{r}
gene_int <- c("CAMKV")

PHIPseq_list$R3_rpk_long %>% 
  filter(Vac_cycle == "day_0") %>% 
  #filter(Outcome =="Good") %>% 
  #filter(grepl("pt_", Sample)) %>% 
  dplyr::filter(Gene == gene_int) %>% 
  ggplot(aes(x = Cancer, y = log(Expression))) +
  geom_boxplot() +
    geom_point(aes(col = Response)) +
        scale_color_manual(values = c(colors_LHB$Response))+
  theme(aspect.ratio = 1) + theme_bw()+
  ggtitle(paste0("Expression of ", gene_int, " gene")) 
  
PHIPseq_list$R3_rpk_long %>% 
  filter(Outcome !="HD") %>% 
  #filter(grepl("pt_", Sample)) %>% 
  dplyr::filter(Gene == gene_int) %>% 
 # left_join(df_Phip_metadata, by = c("Sample" = "strings")) %>% 
  ggplot(aes(x = Vac_cycle, y = log(Expression))) +
  geom_boxplot() +
  geom_line(aes(group = Subject_ID), color = "Grey25", alpha = 0.3) +
  geom_point(aes(col = Response)) +
        scale_color_manual(values = c(colors_LHB$Response))+
  theme(aspect.ratio = 1) + theme_bw()+
  ggtitle(paste0("Expression of ", gene_int, " gene")) +
  facet_wrap(~Outcome)
```

### Multi Box Plots (counts)
```{r}
library(patchwork)  # For combining plots into a grid
jitter_box_1 <- position_jitter(width = 0.2, height = 0)
# Define the genes of interest
gene_int <- c(top_sig_genes_up)

# Define comparisons to be made for the stat_compare_means
comparisons <- list(c("Healthy", "Cancer"))  # Replace with actual group names

plot_list <- list()
# Loop over each gene in the gene_int list
for (gene in gene_int) {
  
  # Filter the data for the current gene
  gene_data <- PHIPseq_list$R3_rpk_long %>%
    mutate(Cancer = recode(Cancer,
                           "HD" = "Healthy",
                           "Melanoma" = "Cancer")) %>% 
    filter(Vac_cycle == "day_0") %>%
    dplyr::filter(Gene == gene)
  gene_data$Cancer <- factor(gene_data$Cancer, levels=c("Healthy","Cancer"))
  # Calculate the maximum y value for the current gene
  max_y <- max(log2(gene_data$Expression+0.01), na.rm = TRUE)
  
  # Create the plot
  p <- gene_data %>%
    ggplot(aes(x = Cancer, y = log2(Expression+0.01))) +
    geom_boxplot(outlier.shape = NA) +
    #geom_point(aes(col = Response)) +
     geom_jitter(position = jitter_box_1, aes(col = Response),
                 size = 2, alpha = 0.5) +
    scale_color_manual(values = c(colors_LHB$Response)) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
    
    theme_prism() +
      labs(title = "Baseline PHIP-seq expression",
           # x = "",
           y = "PHIP-seq log(rpk)") +
    theme(
      aspect.ratio = 2/1.7,
      legend.position = "none",
      plot.title = element_text(size = 15),
      axis.title.x = element_blank(),  # Remove x-axis label
      axis.title.y = element_blank()   # Remove y-axis label
    ) +
    stat_compare_means(
      method = "wilcox.test", 
      comparisons = comparisons,      # List of comparisons to make
      aes(label = sprintf("p = %.2f", as.numeric(..p.format..))),  
      label.y = max_y + 0.5,          
      size = 4.5,                       
      tip.length = 0.02,               
       vjust = 0.05                    
    ) +
    ggtitle(paste0(gene))

  plot_list[[gene]] <- p   # Add the plot to the list
}

# Combine all plots into a grid using patchwork
combined_plot <- wrap_plots(plot_list, ncol = 5) 
print(combined_plot)
# combined_plot + plot_layout(guides = "collect") 
#   theme(legend.position = "right")# & labs(x = "Serveur", y = "Distance Parcourue")
```

### Background Expression Significant Gene Plots 
```{r}
limma_toptab_genes <- limma_toptab %>%  # genes from R3 limma analysis 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% pull(Gene)
head(limma_toptab)
Row_sums_rpk$Background
my_df_plot <- Row_sums_rpk %>% mutate(
  Background =  if_else(Gene %in% clear_0count_genes, "Clear", "Background"),
  Significant_gene = if_else(Gene %in% limma_toptab_genes, "Significant", NA)
) %>% left_join(limma_toptab %>% dplyr::select(Gene,Cancer), by = "Gene")


head(my_df_plot)
my_df_plot %>% 
  ggplot(aes(x = log2(Row_sums_HD), y = log2(Row_sums_Pts))) +
   geom_point(data = my_df_plot %>% filter(Background == "Background") %>% filter(is.na(Cancer)),
             color = "grey", size = 0.5) +  # First set of points for Healthy
  geom_point(data = my_df_plot %>% filter(Background == "Clear") %>% filter(is.na(Cancer)), 
             color = "black", size = 0.5) +  # First set of points for Healthy
  geom_point(data = my_df_plot %>% filter(Background == "Clear") %>% filter(Cancer == "Healthy") %>% 
               filter(Significant_gene == "Significant"),
              color = "#88B8CA", size = 1.5) +  # First set of points for Healthy
  geom_point(data = my_df_plot %>% filter(Background == "Clear") %>% filter(Cancer == "Cancer") %>% 
               filter(Significant_gene == "Significant"),
              color = "#FFB6C1", size = 1.5) +  # Second set of points for Melanoma
  # geom_text(aes(label = ifelse(Gene %in% c(top_sig_genes_up, top_sig_genes_down), Gene, NA)),  # Label significant genes
  #           vjust = 1.5, hjust = 1, check_overlap = TRUE) +
    #geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # 45-degree line
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black", size = 0.5) +  # Correlation line
  #   geom_text_repel(
  #   aes(label = ifelse(Gene %in% c(top_sig_genes_up), Gene, NA)),color = "red4",  # Label significant genes
  #   #vjust = 1.5, hjust = -1,
  #   box.padding = 0.25,     # Padding around labels
  #   point.padding = 0.1,   # Padding around points
  #   segment.color = 'black', # Color of the line segment
  #   max.overlaps = Inf,    # Allow maximum number of overlaps before truncating
  #   na.rm = TRUE           # Remove NA labels
  # ) +
  labs(title = "Melanoma Patients vs Healthy Controls",  
       x = expression("Log" [2] * "HD (rpk)"),
       y = expression("Log" [2] * "Melanoma (rpk)")) +        
  theme_prism()  +
  theme(aspect.ratio = 1/1,
        axis.text.x = element_text(face="plain", color="black",size=16), 
        axis.text.y = element_text(face="plain", color="black",size=16),  
        axis.title.x= element_text(face="plain", color="black",size=16),                        
        axis.title.y= element_text(face="plain", color="black",size=16))
  
```

