---
title: "Autoantidoby manuscript (Vaccine-induced kinetic analyses)"
output: html_document
date: "2024-09-20"
editor_options: 
  chunk_output_type: console
---

- in this script I will apply limma model to determine which auto-abs are enrichned in Melanoma vs HD samples
at baseline.
- I will perform differential gene expression
- visualizations: volcano plots, counts 

# Paths/libraries/source code
```{r}
# ---- Project setup: paths, packages, colours/themes ----------------------
# Replaces the original machine-specific absolute paths and the two lost
# source files (autoantibody_libraries.R, LHB_color_tools.R). setup.R defines
# file_path, data_path, results_path and loads all packages + figure themes.
if (file.exists("setup.R")) source("setup.R") else source(file.path("R", "setup.R"))
```

# read PHIPseq_list
```{r}
PHIPseq_list <-  readRDS(paste0(data_path, "PHIPseq_list.Rda"))
names(PHIPseq_list)
```

## 1. select subjects
```{r}
VAC_CYC = "_0|_43|_89"
SUBJ_TYPE = c("Melanoma")
VAR = "Outcome" 
VAR_VAL = c("Good","Bad")
```
### a) define matrix_phip & meta_phip
### b) filter background genes
```{r}
# Define the function with an additional variable and value for filtering
filter_meta_subjects <- function(meta_name, vac_cycle = NULL, cancer = NULL, outcome = NULL, outcome_val = NULL) {
  # Apply filtering based on input arguments
  
  filtered_df <- PHIPseq_list[[meta_name]] %>%
    filter(
      (is.null(vac_cycle) | grepl(vac_cycle, Vac_cycle)) &
      (is.null(Cancer) | Cancer %in% c(cancer))
    )
  
  # Apply additional filtering if specified by outcome
  if (!is.null(outcome) & !is.null(outcome_val)) {
    filtered_df <- filtered_df %>%
      filter(!!sym(outcome) %in% c(outcome_val))
  }
  
  # Extract and return the 'strings' column
  return(filtered_df %>% pull(strings))
}

# SELECT subjects for analyses
filtered_subjects <- filter_meta_subjects(meta_name = "df_Phip_metadata",
                                           vac_cycle = c(VAC_CYC), 
                                            cancer = c(SUBJ_TYPE), 
                                            outcome = VAR, outcome_val = c(VAR_VAL)
                                          )

# Define a function to subset and format matrix and metadata
prepare_phip_data <- function(data_name, phip_list, pts) {
  # Subset the matrix based on the input dataset name and patients of interest
  matrix_phip <- phip_list[[data_name]] %>% 
    dplyr::select(all_of(intersect(pts, colnames(phip_list[[data_name]]))))
  
  # Subset the metadata to match the columns of the matrix
  meta_phip <- phip_list$df_Phip_metadata %>% 
    filter(strings %in% colnames(matrix_phip))
  
  # Set the row names of the metadata to match the column names of the matrix
  rownames(meta_phip) <- meta_phip$strings
  
  # Return the matrix and metadata as a list
  list(matrix = matrix_phip, metadata = meta_phip)
}

# use of the function and CHOOSE dtaaset
phip_data <- prepare_phip_data("Phip_seq_R3_rpk", PHIPseq_list, filtered_subjects)

# Accessing the resulting matrix and metadata
matrix_phip <- phip_data$matrix
matrix_phip <- matrix_phip[rownames(matrix_phip) %in% PHIPseq_list$clear_0count_genes, ] # remove background genes
meta_phip <- phip_data$metadata
####################
# Check if column names of matrix_phip match row names of meta_phip, if not reorder the names in metadata
if (all(colnames(matrix_phip) == rownames(meta_phip))) {
  print("All matches")
} else if (setequal(colnames(matrix_phip), rownames(meta_phip))) {
  print("The column names are the same but in different order. Reordering meta_phip...")
  meta_phip <- meta_phip[colnames(matrix_phip), ]
  print("meta_phip has been reordered to match column names of matrix_phip.")
} else {
  print("The column names do not match the row names of meta_phip. No action taken.")
}
```

# -----------------------------
# limma model 
# ------------
### 1. Get DE Genes
- resulting table = top_tables_list_filtered
```{r}
library('limma')
####################
meta_phip$Vac_cycle <- factor(meta_phip$Vac_cycle, levels=c("day_0","day_43","day_89_101"))
meta_phip$Outcome <- factor(meta_phip$Outcome, levels=c("Good","Bad"))
meta_phip$Outcome_cycle <- factor(paste(meta_phip$Outcome,meta_phip$Vac_cycle,sep="_")) 
####################

# Outcome vs vaccine cycle design  
design <- model.matrix(~0 + Outcome_cycle, data = meta_phip)
colnames(design) <- levels(meta_phip$Outcome_cycle)

contrast_matrix <- makeContrasts(
  Good_day_43_vs_Good_day_0 = Good_day_43 - Good_day_0,
  Good_day_89_vs_Good_day_0 = Good_day_89_101 - Good_day_0,
  Good_day_89_vs_Good_day_43 = Good_day_89_101 - Good_day_43,
  
  Bad_day_43_vs_Bad_day_0 = Bad_day_43 - Bad_day_0,
  Bad_day_89_vs_Bad_day_0 = Bad_day_89_101 - Bad_day_0,
  Bad_day_89_vs_Bad_day_43 = Bad_day_89_101 - Bad_day_43,
  
   Day_0_Good_vs_Bad = Good_day_0 - Bad_day_0,
   Day_43_Good_vs_Bad = Good_day_43 - Bad_day_43,
   Day_89_Good_vs_Bad = Good_day_89_101 - Bad_day_89_101,

  levels = design
)

# Assuming `matrix_phip` is your data matrix with genes/proteins as rows and samples as columns
# log transformation of the data
log_rpk <- log2(matrix_phip + 0.01)  # Adding 0.01 to avoid log(0)
# Fit the linear model to the log-transformed data
vfit <- lmFit(log_rpk, design)
# Apply the contrast matrix
vfit <- contrasts.fit(vfit, contrast_matrix)
# Perform empirical Bayes smoothing to the standard errors
efit <- eBayes(vfit)
# Extract the top table with all results

top_tables_list <- list()

# Loop through each contrast and extract top tables
for (contrast_name in colnames(contrast_matrix)) {
  # Extract top table for the current contrast
  toptab <- topTable(efit, coef = contrast_name, sort.by = "P", n = Inf)
  
  # Store in the list with the contrast name as the key
  top_tables_list[[contrast_name]] <- toptab
  
  # Print or view the top table for the current contrast
  cat("\nTop table for contrast:", contrast_name, "\n")
  print(head(toptab))  # Displaying the first few rows of the top table
}

# add Gene as a column
top_tables_list <- lapply(top_tables_list, function(x) {
  x$Gene <- rownames(x)
  return(x)
})

# Create a new list to store the filtered data frames
top_tables_list_filtered <- list()
# Loop through each data frame in the top_tables_list
for (tab_name in names(top_tables_list)) {
  # Access the data frame by name
  tab <- top_tables_list[[tab_name]]
  
  # Apply filtering: arrange by logFC and filter based on P.Value and logFC
  filtered_tab <- tab %>% 
    arrange(logFC) %>% 
    filter(P.Value <= 0.05 & abs(logFC) > 0.5)
  
  filtered_tab <- filtered_tab %>% mutate(Direction = if_else(logFC > 0, "Up", "Down")) 
  # Store the filtered data frame in the new list with the same name
  top_tables_list_filtered[[tab_name]] <- filtered_tab
}

lapply(top_tables_list_filtered, function(x) x %>% janitor::tabyl(Direction))
```
# -----------------------------

### Figures 1 (volcano)
- https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
```{r}
# Use gsub to replace underscores and remove everything after '_vs_'
# gsub("_", " ", sub("_vs_.*", "", names(top_tables_list_filtered[1])))
# gsub("_", " ", sub(".*_vs_", "", names(top_tables_list_filtered[1])))

names(top_tables_list_filtered)

# Function to extract part of the string before or after _vs_:
# BEST works with 1:6 - the names later are shorter and do not fit the naming formula (will fix later)
extract_part <- function(input_string, choice = "1st") {
  if (choice == "1st") {
    # Extract the part 1st '_vs_'
    result <- gsub("_", " ", sub("_vs_.*", "", input_string))
  } else if (choice == "2nd") {
    # Extract the part 2nd '_vs_'
    result <- gsub("_", " ", sub(".*_vs_", "", input_string))
  } else {
    stop("Invalid choice. Please use '1st' or '2nd'.")
  }
  
  return(result)
}

names(top_tables_list_filtered[1])
data_set = 1
# Gene expression table
fin.volcano.tab <- data.frame(top_tables_list_filtered[[data_set]]) %>% arrange(desc(logFC))
rownames(fin.volcano.tab) <- fin.volcano.tab$Gene
# up in Melanoma
top_sig_genes_up <- fin.volcano.tab %>% filter(logFC > 0 ) %>% 
  #filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)
# up in HD
top_sig_genes_down <- fin.volcano.tab %>% filter(logFC < 0 ) %>% 
  # filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)

# Volcano Plot
###########

#FIX the labels 
EnhancedVolcano::EnhancedVolcano(top_tables_list[[data_set]],
    lab = rownames(top_tables_list[[data_set]]),
    title = "Baseline comparison",
    subtitle = NULL,
    caption = NULL,
    x = 'logFC', maxoverlapsConnectors = Inf,
    y = 'P.Value',
    selectLab = c(top_sig_genes_up,top_sig_genes_down),
    xlab = bquote(~Log[2]~ 'fold change'),
    ylim = c(0, 8),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 3.0,
    labSize = 4.0,
    labCol = 'black',
    labFace = 'plain',
    boxedLabels = TRUE,
    parseLabels = TRUE,
    col = c('black','grey', 'purple', 'red3'),  # Adjust colors if necessary
    colAlpha = 4/5,
    legendLabels = c('Not significant','LogFC > 0.5', 'p-value <= 0.05', 'p-value <= 0.05 & LogFC > 0.5'),
    legendPosition = 'right',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'black') +
    annotate("text", x = c(-3.5, 4.3), y = 8, label = c(extract_part(names(top_tables_list_filtered[data_set]), "1st"),
                                                        extract_part(names(top_tables_list_filtered[data_set]), "2nd")), size = 8)
```

### Figures 1 (counts)
```{r}
gene_int <- c("TDP2", "PSMA5")
  
PHIPseq_list$R3_rpk_long %>% 
  filter(Outcome !="HD") %>% 
  #filter(grepl("pt_", Sample)) %>% 
  dplyr::filter(Gene %in% c(gene_int)) %>% 
 # left_join(df_Phip_metadata, by = c("Sample" = "strings")) %>% 
  ggplot(aes(x = Vac_cycle, y = log(Expression))) +
  geom_boxplot() +
  geom_line(aes(group = Subject_ID), color = "Grey25", alpha = 0.3) +
  geom_point(aes(col = Response)) +
        scale_color_manual(values = c(colors_LHB$Response))+
  theme(aspect.ratio = 1) + theme_bw()+
  ggtitle(paste0("Expression of ", gene_int, " gene")) +
  facet_wrap(~Outcome+Gene)
```
### Multi Box Plots (counts)
```{r}
library(patchwork)  # For combining plots into a grid
jitter_box_1 <- position_jitter(width = 0.2, height = 0)
# Define the genes of interest
gene_int <- c(top_sig_genes_up)

# Define comparisons to be made for the stat_compare_means
comparisons <- list(c("Healthy", "Cancer"))  # Replace with actual group names

plot_list <- list()
# Loop over each gene in the gene_int list
for (gene in gene_int) {
  
  # Filter the data for the current gene
  gene_data <- PHIPseq_list$R3_rpk_long %>%
    mutate(Cancer = recode(Cancer,
                           "HD" = "Healthy",
                           "Melanoma" = "Cancer")) %>% 
    filter(Vac_cycle == "day_0") %>%
    dplyr::filter(Gene == gene)
  gene_data$Cancer <- factor(gene_data$Cancer, levels=c("Healthy","Cancer"))
  # Calculate the maximum y value for the current gene
  max_y <- max(log2(gene_data$Expression+0.01), na.rm = TRUE)
  
  # Create the plot
  p <- gene_data %>%
    ggplot(aes(x = Cancer, y = log2(Expression+0.01))) +
    geom_boxplot(outlier.shape = NA) +
    #geom_point(aes(col = Response)) +
     geom_jitter(position = jitter_box_1, aes(col = Response),
                 size = 2, alpha = 0.5) +
    scale_color_manual(values = c(colors_LHB$Response)) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
    
    theme_prism() +
      labs(title = "Baseline PHIP-seq expression",
           # x = "",
           y = "PHIP-seq log(rpk)") +
    theme(
      aspect.ratio = 2/1.7,
      legend.position = "none",
      plot.title = element_text(size = 15),
      axis.title.x = element_blank(),  # Remove x-axis label
      axis.title.y = element_blank()   # Remove y-axis label
    ) +
    stat_compare_means(
      method = "wilcox.test", 
      comparisons = comparisons,      # List of comparisons to make
      aes(label = sprintf("p = %.2f", as.numeric(..p.format..))),  
      label.y = max_y + 0.5,          
      size = 4.5,                       
      tip.length = 0.02,               
       vjust = 0.05                    
    ) +
    ggtitle(paste0(gene))

  plot_list[[gene]] <- p   # Add the plot to the list
}

# Combine all plots into a grid using patchwork
combined_plot <- wrap_plots(plot_list, ncol = 5) 
print(combined_plot)
# combined_plot + plot_layout(guides = "collect") 
#   theme(legend.position = "right")# & labs(x = "Serveur", y = "Distance Parcourue")
```

# -----------------------------
# 1. degPatterns analysis
## a) subset patent groups of interest
```{r}
sub_top_list_filt <- top_tables_list_filtered[
  grepl("Good|Bad", names(top_tables_list_filtered))]
names(sub_top_list_filt)
lapply(sub_top_list_filt, head)
```

## b) subset signif DE genes
```{r}
all_DE_genes <- c()
# Iterate over each element in the top_tables_list_filtered
for (i in seq_along(sub_top_list_filt)) {
  # Extract the 'Gene' column from the i-th element of the list and add to the all_genes vector
  all_DE_genes <- c(all_DE_genes, sub_top_list_filt[[i]] %>% pull(Gene))
  all_DE_genes <- unique(all_DE_genes)
}
length(all_DE_genes)
```

## c) run DEGreport
 - clusters_vacc.cycle_outcome was previoulsy run and saved into PHIPseq_list
 - we can recall it (no need to run this algorythm)
```{}
library('DEGreport')
# selected matrices from above code lines:
#matrix_phip
#meta_phip

if(exists("matrix_phip") & exists("meta_phip")){
  print("Both tables are present")
}
colnames(matrix_phip)  == rownames(meta_phip)

sel_matrix_phip <- matrix_phip[rownames(matrix_phip) %in% all_DE_genes, ]
colnames(sel_matrix_phip)

# matrix has to be log transformed
res_10 <- degPatterns(log2(sel_matrix_phip+0.01), meta_phip, time="Vac_cycle", col = "Outcome", 
                   reduce = TRUE,
                   minc = 25,
                   nClusters = 12)
res_10
res_10$df %>% group_by(cluster) %>% count() %>% arrange(desc(n)) %>% filter(n > 100)

clust.tab <- res_10[["normalized"]]
clust.tab %<>% dplyr::select(genes:value, -merge, Vac_cycle,Outcome,Outcome_cycle,cluster)
# clust.tab %>% 
#   mutate(cluster = if_else(cluster %in% c("8","10"), "8",cluster ))
```

### - save to PHIPseq_list
```{r}
#PHIPseq_list$clusters_vacc.cycle_outcome <- clust.tab
#PHIPseq_list %>% saveRDS(paste0(data_path, "PHIPseq_list.Rda"))

clust.tab <- PHIPseq_list$clusters_vacc.cycle_outcome
clust.tab$cluster <- as.factor(clust.tab$cluster) 
clust.tab %<>% mutate(Vac_cycle = gsub("day_","",Vac_cycle),
                      Vac_cycle = gsub("89_","",Vac_cycle))
clust.tab$Vac_cycle <- factor(clust.tab$Vac_cycle, levels=c("0","43","101"))
```
## d) plot Deg Patterns
```{r}
cluster_labels <- setNames(paste("Cluster", unique(clust.tab$cluster)), unique(clust.tab$cluster))

ggplot(clust.tab, aes(Vac_cycle, value, color = Outcome)) +
  geom_line(data = clust.tab %>% filter(Outcome == "Good"),
            aes(group = genes),
            color = "red", size = 0.1, alpha = 0.5) +  
  geom_line(data = clust.tab %>% filter(Outcome == "Bad"), 
            aes(group = genes),
            color = "grey45", size = 0.1, alpha = 0.5) +  
  scale_color_manual(values = c(colors_LHB$Outcome)) +
  theme_prism() +
  # geom_smooth(aes(group = Outcome), method = "auto") + 
   scale_color_manual(values = c(colors_LHB$Out_median_lines)) +
  theme_prism() +
  geom_smooth(aes(group = Outcome), method = "loess") + 
  labs(title = "Auto-antibody Expression Profiles",
              #subtitle = "adj.p < 0.05",
              x = "Vaccine cycle",                         
       y = "Scaled(rpk)")  +
    theme(
    #plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 14,angle = 0, face = "bold"),
    strip.background =element_rect(fill="white", color = NA),
    strip.placement = "outside",
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    #axis.text = element_text(size = 14),
    legend.position = "right",
    panel.grid.major.y = element_blank()) +
  facet_wrap(~cluster, labeller = labeller(cluster = cluster_labels)) 
```

```{r}
cluster_df <- clust.tab %>% 
  dplyr::select(-Outcome_cycle,-Outcome, -value,- Vac_cycle) %>% distinct() %>% 
  arrange(cluster) %>% 
  mutate(ENTREZID = mapIds(org.Hs.eg.db, keys = genes, keytype="SYMBOL", column = "ENTREZID")) 

cluster_df %>% group_by(cluster) %>% count() 
cluster_list <- cluster_df %>%          
  group_by(cluster) %>%                 
  group_split(.keep = TRUE)             

names(cluster_list) <- unique(cluster_df$cluster)
```

# PATHWAY ENRICHMENT
### - libraries
```{r}
library(clusterProfiler)
library(GSEABase)
library(msigdbr)
library(fgsea)
library(enrichplot)
library(org.Hs.eg.db)
```

## 1. ReactomePA
```{}
library('ReactomePA')

geneUniverse <- mapIds(org.Hs.eg.db, keys = rownames(matrix_phip), keytype="SYMBOL", column = "ENTREZID")
geneUniverse <- geneUniverse[!is.na(geneUniverse)]


# Reactome pathway over-representation analysis
perform_enrichment <- function(cluster_data) {
  #set.seed(1234)
  # Extract gene list from the cluster data (assuming 'ENTREZID' is the gene column)
  gene_list <- unname(cluster_data$ENTREZID)
  enrich_result <- enrichPathway(gene = gene_list, 
                                 pvalueCutoff = 0.05, 
                                 pAdjustMethod = "BH", 
                                 qvalueCutoff = 0.5,
                                 minGSSize = 10, 
                                 maxGSSize = 5000,
                                 readable = TRUE)
  return(enrich_result@result)
}

# Apply the function to each cluster in the list
enrichPathway_results <- lapply(cluster_list, perform_enrichment)

lapply(enrichPathway_results, function(x) x %>% filter(pvalue <= 0.05) %>% nrow())
```

## 2. enrichGO
```{}
perform_enrichGO <- function(cluster_data) {
  # Extract gene list from the cluster data (assuming 'ENTREZID' is the gene column)
  gene_list <- unname(cluster_data$ENTREZID)
  
  # Perform enrichment analysis using enrichPathway
  enrich_result <- enrichGO(gene = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENTREZID',
                ont           = "MF", #“BP”, “MF”, “CC”, “ALL”
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05)
  
  return(enrich_result@result)
}

# Apply the function to each cluster in the list
enrichGO_results <- lapply(cluster_list, perform_enrichGO)

enrichGO_results[[1]] %>% names()
lapply(enrichGO_results, function(x) x %>% filter(pvalue <= 0.05) %>% nrow())
lapply(enrichGO_results, function(x) x %>% filter(qvalue <= 0.05) %>% nrow())

enrichGO_results[[1]] %>%
   arrange(desc(Count)) %>% filter(pvalue <= 0.05) %>% 
  filter(Count >= 3) %>% 
  # head(25) %>%
  ggplot(aes(x = -log2(pvalue), y =  fct_reorder(Description, -log2(pvalue)))) + #geom_col(aes(fill = DF_label)) + 
  #geom_segment(aes(yend=Description), xend=0, colour="grey50", linewidth=1, ) +
  geom_point(aes(size = Count, color = as.factor(Count))) +
    theme_bw() +
  labs(title = "Pathway Enrichment",
              #subtitle = "adj.p < 0.05",
              x = "-log2(pvalue)",                         
       y = "Pathway")  +
   scale_size_continuous(range = c(0, 9)) +
  theme(strip.text.y = element_blank()) +
    facet_grid(rows = vars(Count), scales = "free_y", #switch = "y", 
             space = "free_y") 
```

## 3. enrichWP
```{r}
perform_enrichWP <- function(cluster_data) {
  # Extract gene list from the cluster data (assuming 'ENTREZID' is the gene column)
  gene_list <- unname(cluster_data$ENTREZID)
  
  # Perform enrichment analysis using enrichPathway
  enrich_result <- enrichWP(gene = gene_list,
                            organism = "Homo sapiens")
  
  return(enrich_result@result)
}

# Apply the function to each cluster in the list
enrichWP_results <- lapply(cluster_list, perform_enrichWP)

enrichWP_results[[1]] %>% names()
lapply(enrichWP_results, function(x) x %>% filter(pvalue <= 0.05) %>% nrow())

enrichWP_results[[1]] %>%
   arrange(desc(Count)) %>% filter(pvalue <= 0.05) %>% 
  filter(Count >= 3) %>% 
  # head(25) %>%
  ggplot(aes(x = -log2(pvalue), y =  fct_reorder(Description, -log2(pvalue)))) + #geom_col(aes(fill = DF_label)) + 
  #geom_segment(aes(yend=Description), xend=0, colour="grey50", linewidth=1, ) +
  geom_point(aes(size = Count, color = as.factor(Count))) +
    theme_bw() +
  labs(title = "Pathway Enrichment",
              #subtitle = "adj.p < 0.05",
              x = "-log2(pvalue)",                         
       y = "Pathway")  +
   scale_size_continuous(range = c(0, 9)) +
  theme(strip.text.y = element_blank()) +
    facet_grid(rows = vars(Count), scales = "free_y", #switch = "y", 
             space = "free_y")
```

# ============
# $ save/read Pathway_res_list
```{r}
Pathway_res_list <- list()
Pathway_res_list$enrichPathway_res <- enrichPathway_results
Pathway_res_list$enrichGO_res <- enrichGO_results
Pathway_res_list$enrichWP_res <- enrichWP_results

#Pathway_res_list %>%  saveRDS("Pathway_results_list.Rda")
Pathway_res_list <-  readRDS(file.path(data_path, "Pathway_results_list.Rda"))
names(Pathway_res_list)

# Combine the data frames and add an ID column based on the element index
combined_df <- do.call(rbind, lapply(1:12, function(i) {
  df <- Pathway_res_list[[1]][[i]]
  df$Cluster <- paste0(i)  # Add a column indicating the data frame's index
  #df$Count <- as.factor(df$Count)  # Add a column indicating the data frame's index
  return(df)
}))

glimpse(combined_df)

combined_df %<>% mutate(geneID_split = gsub("/", " ", geneID))
```
### $ selecet a cluster
```{r}
result_list <- lapply(Pathway_res_list[1:3], function(x) x[[6]])
result_list[[3]] %>% names()
# Bind the extracted data frames into a single data frame
combined_df <- do.call(rbind, result_list)
```
### $ functions
```{r}
# Function to add a newline after every 10th gene
add_newline_after_10 <- function(gene_string) {
  genes <- unlist(strsplit(gene_string, " "))  # Split by space
  if (length(genes) > 10) {
    # Insert newline after every 10th gene
    paste(sapply(seq(1, length(genes), by = 10), function(i) {
      paste(genes[i:min(i+9, length(genes))], collapse = " ")
    }), collapse = "\n")
  } else {
    gene_string  # If <= 10, return the original string
  }
}

add_newline_after_5 <- function(gene_string) {
  genes <- unlist(strsplit(gene_string, " "))  # Split by space
  if (length(genes) > 3) {
    # Insert newline after every 10th gene
    paste(sapply(seq(1, length(genes), by = 3), function(i) {
      paste(genes[i:min(i+9, length(genes))], collapse = " ")
    }), collapse = "\n")
  } else {
    gene_string  # If <= 10, return the original string
  }
}
```
### $ CLUST gene selections
```{r}
Gene_count <- 2
######### CLUST_6 #########
CLUST_6_filters <- c("Signaling by EGFR in Cancer",
                            "Interleukin-3, Interleukin-5 and GM-CSF signaling",
                            "Signaling by EGFR in Cancer",
                            "Integrin signaling",
                            "Initial triggering of complement",
                            "Response of Mtb to phagocytosis",
                            "Transcriptional regulation of brown and beige adipocyte differentiation",
                            "DAP12 signaling",
                            "TP53 Regulates Transcription of Cell Cycle Genes",
                            "Signaling by ALK in cancer",
                            "Sialic acid metabolism",
                            "Costimulation by the CD28 family",
                            "Disease",
                            "Hemostasis",
                            "Ion channel transport",
                            "Vesicle-mediated transport",
                            "Platelet activation, signaling and aggregation",
                            "Signaling by CSF1 (M-CSF) in myeloid cells")

######### CLUST_2 #########
CLUST_2_filters <- c("Glycogen synthesis|storage diseases|glycogenolysis|trafficking and regulation|L1|Processing of Intronless Pre-mRNAs|Intronless Pre-mRNA|RNA Polymerase II")

CLUST_2_filters <- combined_df %>%
  filter(Cluster == 6) %>% 
  filter(pvalue <= 0.05) %>% 
  filter(Count >= Gene_count) %>%
  filter(!grepl(CLUST_2_filters,Description)) %>% pull(Description)

######### CLUST_9 #########
CLUST_9_filters <- c("RAF-independent MAPK1/3|Xenobiotics|Gastrulation|Synthesis of bile acids and bile salts|mutants|Paradoxical|Synaptic adhesion-like|neural plate|Negative regulation of MAPK|Fatty acids|Ras activation upon Ca2+|RAF/MAP kinase cascade|fusions|RHOBTB2|RHOBTB1|GABA synthesis|MAPK1/MAPK3 signaling|CYP2E1 reactions|Dopamine Neurotransmitter|MAPK3|Serotonin Neurotransmitter|IL-6-type cytokine receptor|KIT in disease|Norepinephrine|Glycosphingolipid|Signalling to ERKs|RHOBTB")

CLUST_9_filters <- combined_df %>%
  filter(Cluster == 9) %>% 
  filter(pvalue <= 0.05) %>% 
  filter(Count >= Gene_count) %>%
  filter(!grepl(CLUST_9_filters,Description)) %>% pull(Description)

######### CLUST_10 #########

CLUST_10_filters <- c("UCH proteinases",
                     "Metabolism of lipids",
                     "Neddylation",
                     "Class I MHC mediated antigen processing & presentation",
                     "activated TAK1 mediates p38 MAPK activation",
                     "p38MAPK events",
                     #"Olfactory Signaling Pathway",
                     "Innate Immune System",
                     "Neutrophil degranulation",
                     "Signaling by Hedgehog",
                     "APC/C-mediated degradation of cell cycle proteins",
                     "TCR signaling",
                     "C-type lectin receptors (CLRs)",
                     "Beta-catenin independent WNT signaling",
                     "Cross-presentation of soluble exogenous antigens (endosomes)",
                     "Interleukin-1 family signaling",
                     "Disorders of transmembrane transporters",
                     "SLC transporter disorders",
                     "Retrograde transport at the Trans-Golgi-Network")
######### CLUST_8 #########
CLUST_8_filters <- c("Neutrophil degranulation",
                      "Cytoprotection by HMOX1",
                      "Cellular response to chemical stress",
                      "RHOBTB2 GTPase cycle",
                      "Formation of tubulin folding intermediates by CCT/TriC")
######### CLUST_5 #########
CLUST_5_filters <- c("Mitochondrial Fatty Acid Beta-Oxidation",
                      "Downregulation of TGF-beta receptor signaling",
                     "TP53 Regulates Transcription of Genes Involved in Cytochrome C Release",
                     "Cytokine Signaling in Immune system",
                     "Citric acid cycle (TCA cycle)",
                     "Interleukin-12 signaling",
                     "Metabolism of RNA",
                     "Metabolism of amino acids and derivatives"
                     )
######### CLUST_11 #########
CLUST_11_filters <- c("Biological oxidations",
                     "MITF-M-regulated melanocyte development",
                     "MAPK6/MAPK4 signaling",
                     "PTEN Regulation",
                     "Immune System",
                     "Innate Immune System",
                     "Regulation of actin dynamics for phagocytic cup formation",
                     "Metabolism",
                     "Fcgamma receptor (FCGR) dependent phagocytosis",
                     "Pyroptosis",
                     "Programmed Cell Death",
                     "Neutrophil degranulation",
                     "Hemostasis",
                     "Cytokine Signaling in Immune system",
                     "Signaling by VEGF",
                     "DDX58/IFIH1-mediated induction of interferon-alpha/beta",
                     "Chaperone Mediated Autophagy"
                     )
######### CLUST_7 #########
CLUST_7_filters <- c(
  "Metabolism of lipids",
  "Signal Transduction",
  "Complement cascade",
  "MITF-M-regulated melanocyte development",
  "Regulation of MITF-M-dependent genes involved in pigmentation",
  "Activation of Matrix Metalloproteinases",
  "Cytosolic sensors of pathogen-associated DNA",
  "Oncogenic MAPK signaling"
)
######### CLUST_3 ######### 
CLUST_3_filters <- c(
  "Oncogenic MAPK signaling",
  "ABC transporters in lipid homeostasis",
  "PPARA activates gene expression",
  "Adipogenesis",
  "ADORA2B mediated anti-inflammatory cytokines production",
  "NOTCH1 Intracellular Domain Regulates Transcription",
  "Parasitic Infection Pathways",
  "Signaling by Nuclear Receptors",
  "Gene expression (Transcription)",
  "mRNA Splicing",
  "Processing of Capped Intron-Containing Pre-mRNA"
)
#########


```
## 3. Dot Plot Cluster Pathways
```{r}
CLUST <- 3
Gene_count <- 2
filter_pathwas <- CLUST_3_filters
# Create the plot
head(combined_df)
combined_df %>%
  filter(Cluster == CLUST) %>% 
  filter(pvalue <= 0.05) %>% 
  filter(Count >= Gene_count) %>%
 filter(Description %in% c(filter_pathwas)) %>%
  ggplot(aes(x = -log2(pvalue), y = fct_reorder(Description, -log2(pvalue)))) +
  geom_point(aes(size = Count), color = "navy", alpha = 0.8) +  # Keep size aesthetic here
  labs(title = paste0("Pathway Enrichment\nCluster: ", CLUST),
       x = "-log2(pvalue)",
       y = "Pathway",
       size = "Gene Count") +  # Correct the label for 'size'
  scale_size_continuous(range = c(0, 9)) +
  theme_classic() + 
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    strip.text = element_text(size = 14, angle = 0, face = "bold"),
    strip.background = element_rect(fill = "white", color = NA),
    strip.placement = "outside",
    axis.line = element_line(size = 1),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold"),
    legend.position = "right",
    panel.grid.major.y = element_blank()
  ) + 
  theme(
     legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14, face = "bold"), 
    legend.key.size = unit(1, "cm"), 
    legend.key.height = unit(0.5, "cm")
  )
####################################################################################
```

## 4. Table Genes in Pathways
```{r}
head(combined_df)
df <- combined_df %>%
  filter(Cluster == CLUST) %>%
    filter(pvalue <= 0.05) %>% 
  filter(Count >= Gene_count) %>%
  arrange(pvalue) %>%
  filter(Description %in% c(filter_pathwas)) %>%
  dplyr::select(Pathway = Description, `gene ID` = geneID_split) %>%
  mutate(`gene ID` = sapply(`gene ID`, add_newline_after_10))

library('flextable')
ft <- flextable(df)
ft <- width(ft, width = 5)
ft <- bold(ft, j = 1, bold = TRUE)
ft <- italic(ft, j = 2, italic = TRUE)
ft <- bold(ft, part = "header", bold = TRUE) # Make the first row bold
ft <- fontsize(ft,part = "header",  size = 13) 

ft <- bg(ft, part = "header", bg = "grey") # Set the background color of the header row to grey
ft <- fontsize(ft, j = 1, size = 14) # Change the font size of the first column to 10
ft <- fontsize(ft, j = 2, size = 10)
ft
save_as_image(ft, path = file.path(fig_path, "flextable.png"
                                #,res = 200 # to adjust resolution if needed
                                ))
```

## 5. Selected Pathway Highlights
```{r}
# Apply the transformation to the `gene ID` column for the specific pathway
path_genes <- df %>%
  filter(grepl("Disease", Pathway)) %>%
  pull(`gene ID`) %>%
  as.character() %>%
  strsplit(" ") %>%    # Split by space
  unlist()

plot.tab <- clust.tab %>% filter(cluster == CLUST)

ggplot(plot.tab, aes(Vac_cycle, value, color = Outcome)) +
  geom_line(data = plot.tab %>% filter(Outcome == "Good"),
            aes(group = genes),
            color = "red", size = 0.1, alpha = 0.5) +  
  geom_line(data = plot.tab %>% filter(Outcome == "Bad"), 
            aes(group = genes),
            color = "grey45", size = 0.1, alpha = 0.5) +  
  # geom_line(data = clust.tab %>% filter(cluster == CLUST) %>% filter(Outcome == "Good") %>%
  #             filter(genes %in% c(path_genes)),
  #           aes(group = genes),
  #           color = "magenta", size = 0.5, alpha = 0.5) +
  scale_color_manual(values = c(colors_LHB$Out_median_lines)) +
  theme_prism() +
  geom_smooth(aes(group = Outcome), method = "loess") + 
  labs(title = paste0("Cluster ", CLUST ," Expression Profile"),
              x = "Vaccine cycle",                         
       y = "Scaled(rpk)")  +
    theme(
    #plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 14,angle = 0, face = "bold"),
    strip.background =element_rect(fill="white", color = NA),
    strip.placement = "outside",
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    #axis.text = element_text(size = 14),
    legend.position = "right",
    panel.grid.major.y = element_blank())
```






# =====
### sorted vector GSEA
## 4. KEGG
```{r}
mkk2 <- gseMKEGG(geneList = cluster_list[[9]]$ENTREZID, # needs to be a sorted vector
                 organism = 'hsa',
                 pvalueCutoff = 1)
head(mkk2)
```

## 5. enrichGO 
```{r}
ja_vec <- top_tables_list_filtered$Day_89_Good_vs_Bad %>%  
  dplyr::select(Gene,logFC) %>% 
  arrange(desc(logFC)) %>% 
  mutate(ENTREZID = mapIds(org.Hs.eg.db, keys = Gene, keytype="SYMBOL", column = "ENTREZID")) %>% 
  pull(ENTREZID,logFC)
ja_vec <- ja_vec[!is.na(ja_vec)]


geneUniverse <- mapIds(org.Hs.eg.db, keys = rownames(matrix_phip), keytype="SYMBOL", column = "ENTREZID")
geneUniverse <- geneUniverse[!is.na(geneUniverse)]

# Run GO enrichment analysis
ego <- enrichGO(
    gene = ja_vec,
    OrgDb = "org.Hs.eg.db",
    keyType = "ENTREZID",
    ont = "BP",
    universe = geneUniverse,
    pvalueCutoff=0.05,
    pAdjustMethod="BH", 
    qvalueCutoff=0.05,
    #readable = TRUE
)
ego_summary <- slot(ego, "result") %>%
    as_tibble() 
head(ego_summary)
ego_summary %>% arrange(Count) %>% filter(pvalue <= 0.05)
hist(ego_summary$FoldEnrichment)
ego_summary %>% 
  filter(Count >= 10) %>% 
   arrange(p.adjust) %>% filter(pvalue <= 0.05) %>% arrange(FoldEnrichment) %>% 
 # head(35) %>%
  #filter(grepl("TNF|NECROSIS|NFK|KAPP|MEK",ID)) %>% 
  #group_by(DF_label) %>% slice(1:35) %>% 
  ggplot(aes(x = FoldEnrichment, y =  fct_reorder(Description, FoldEnrichment))) + #geom_col(aes(fill = DF_label)) + 
  geom_segment(aes(yend=Description,), xend=0, colour="grey50", linewidth=1, ) +
  geom_point(aes(size = Count, fill = pvalue)) +
  scale_color_brewer(palette = "Dark2") 
  facet_grid(rows = vars(DF_label), scales = "free_y", #switch = "y", 
             space = "free_y") +
    theme_bw() +
   labs(title = "Top TNF-TNFR/NFkB Pathways at 4hr",
              subtitle = "adj.p < 0.05")  + xlab("NES")+ ylab("GSEA PAthways") 
```

## 6. a) msigdbr GSEA 
```{r}
# load in gene sets
all_gene_sets = msigdbr(species = "Homo sapiens")
all_gene_sets %<>% filter(gs_subcat %in% 
                            c("CP:BIOCARTA","GO:BP","GO:CC","GO:MF",
                              "CP:KEGG","CP:PID", "CP:REACTOME","CP:WIKIPATHWAYS",
                              "HPO","CP")) # filter out MIRNA gene sets

all_gene_sets$gs_subcat %>% unique()
#all_gene_sets %>% filter(gs_subcat == "TFT:GTRD") %>% pull(gs_name) %>% unique()

# we can also subset libraries
# mm_currated_sets <- msigdbr(
#   species = "Homo sapiens", # Replace with species name relevant to your data
#   category = "H"
# )
```

## 6. b) GSEA 
```{r}
ja_vec_sumb <- top_tables_list_filtered$Day_89_Good_vs_Bad %>%  
  dplyr::select(Gene,logFC) %>% 
  arrange(desc(logFC)) %>% 
  #mutate(ENTREZID = mapIds(org.Hs.eg.db, keys = Gene, keytype="SYMBOL", column = "ENTREZID")) %>% 
  pull(logFC,Gene)

set.seed(123)
out_gsea <- GSEA(
  geneList = ja_vec_sumb, # Ordered ranked gene list
  minGSSize = 5, # Minimum gene set size
  maxGSSize = 5000, # Maximum gene set set
  pvalueCutoff = 1, # p-value cutoff
   pAdjustMethod = "BH", # Benjamini-Hochberg correction
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  TERM2GENE = dplyr::select(
    all_gene_sets,
    gs_name,
    gene_symbol
  )
)

# This code gives overlapping genes from the leading gene set
###############################################################
genes_for_overlap <- out_gsea@result %>%   
    filter(core_enrichment >= 10) %>% 
 arrange(pvalue) %>% filter(pvalue <= 0.05) %>% filter(abs(NES) > 1.75) %>% 
  pull(core_enrichment)

combined_genes <- paste(genes_for_overlap, collapse = "/")
split_gene_sets <- strsplit(combined_genes, "/")

# Find the intersection (overlap) across all gene sets
overlap_genes <- unique(Reduce(intersect, split_gene_sets))
# Output the overlapping genes
print(overlap_genes)
###############################################################

out_gsea@result$gene_number <- sapply(strsplit(out_gsea$core_enrichment, "/"), length)

out_gsea@result %>%   
    filter(core_enrichment >= 10) %>% 
 arrange(pvalue) %>% filter(pvalue <= 0.05) %>% filter(abs(NES) > 1.75) %>% 
  ggplot(aes(x = NES, y =  fct_reorder(ID, NES))) + #geom_col(aes(fill = DF_label)) + 
  geom_segment(aes(yend=ID), xend=0, colour="grey50", linewidth=1, ) +
  geom_point(aes(size = gene_number)) +
  scale_color_brewer(palette = "Dark2") +
    theme_bw() +
   labs(title = "Top TNF-TNFR/NFkB Pathways at 4hr",
              subtitle = "adj.p < 0.05")  + xlab("NES")+ ylab("GSEA PAthways") +
  theme(
    #plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
    plot.title = element_text(size = 15, face = "bold"),
    strip.text.y = element_text(angle = 0, face = "bold"),
             strip.background =element_rect(fill="white", color = NA),
    strip.placement = "outside",
    axis.title.x = element_text(margin = margin(t = 0.5, b = 0.5, unit = "cm")),
    axis.title.y = element_blank(),
    axis.text = element_text(size = 10),
    legend.position = "right",
    panel.grid.major.y = element_blank())


  # facet_grid(rows = vars(DF_label), scales = "free_y", #switch = "y", 
  #            space = "free_y") +

out_gsea@result %>%   
    filter(core_enrichment >= 10) %>% 
    arrange(pvalue) %>% 
    filter(pvalue <= 0.05) %>% 
    filter(abs(NES) > 1.75) %>% 
    ggplot(aes(x = NES, y = fct_reorder(ID, NES))) + 
    geom_segment(aes(yend = ID), xend = 0, colour = "grey50", linewidth = 1) +
  xlim(-2,10)+
    geom_point(aes(size = gene_number)) +
    geom_text(aes(label = core_enrichment), hjust = 0, vjust = 0, size = 2) +
    scale_color_brewer(palette = "Dark2") +
    theme_bw()

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)

# Create the plot
gsea_plot <- out_gsea@result %>%   
    filter(core_enrichment >= 10) %>% 
    arrange(pvalue) %>% 
    filter(pvalue <= 0.05) %>% 
    filter(abs(NES) > 1.75) %>% 
    ggplot(aes(x = NES, y = fct_reorder(ID, NES))) + 
    geom_segment(aes(yend = ID), xend = 0, colour = "grey50", linewidth = 1) +
    geom_point(aes(size = gene_number)) +
    scale_color_brewer(palette = "Dark2") +
    theme_bw()

# Create the table
gsea_table <- out_gsea@result %>% 
    filter(core_enrichment >= 10) %>% 
    arrange(pvalue) %>% 
    filter(pvalue <= 0.05) %>% 
    filter(abs(NES) > 1.75) %>% 
    dplyr::select(filtered_gene_sets) 

# Create the table (we'll use the filtered_gene_sets and IDs from the filtered data)
gsea_table <- filtered_data %>% 
    dplyr::select(ID, filtered_gene_sets)

# Convert the table to grob, with the row names (IDs) as the first column
table_grob <- tableGrob(gsea_table, rows = NULL, theme = ttheme_minimal(
  core = list(padding = unit(c(5, 5), "mm")),
  colhead = list(padding = unit(c(5, 5), "mm"))
))

# Arrange the plot and the table side by side
grid.arrange(gsea_plot, table_grob, ncol = 2, widths = c(2/3, 1/3))

```
# =====

```{r}
# Vector of genes to filter
filter_genes <- unlist(split_gene_sets)

# Define a function to filter each gene set
filter_gene_set <- function(core_enrichment, filter_genes) {
  # Split the gene set by "/"
  genes <- unlist(strsplit(core_enrichment, "/"))
  
  # Filter only the genes present in the filter_genes vector
  filtered_genes <- genes[!genes %in% filter_genes]
  
  # Combine back into a single string
  paste(filtered_genes, collapse = "/")
}

# Apply the function to each row of gene sets
out_gsea@result$filtered_gene_sets <- sapply(out_gsea@result$core_enrichment, filter_gene_set, 
                                             filter_genes = filter_genes)

# Output the updated data frame
print(out_gsea@result$filtered_gene_sets)

```

```{r}
ego_summary %>% arrange(pvalue)

clust.tab %>% filter(Vac_cycle)

abc_wide <- clust.tab %>%  dplyr::select(genes, Outcome,cluster, value) %>% 
  pivot_wider(names_from = Outcome, values_from = value)
abc_wide %<>% mutate(diff = Good - Bad) %>% 
  mutate(up_down = if_else(diff < 0, "down", "up"))

abc_wide %>% arrange(desc(diff)) %>% group_by(cluster) %>% count() %>% arrange(desc(n))
abc_wide %>% arrange(desc(diff)) %>% group_by(cluster,up_down) %>% count() %>% filter(n > 25) %>% 
  ggplot(aes(up_down, n)) +  geom_bar(stat = "identity", aes(fill = up_down)) +
  facet_wrap(~cluster)




abc <- clust.tab %>% filter(Vac_cycle == "day_43")
names(clust.tab)

abc %>% ggplot(aes(Outcome, value)) +
  geom_point(aes(color = Outcome)) + 
   geom_line(aes(group = genes),
             color = "red", size = 0.1, alpha = 0.5)+
       scale_color_manual(values = c(colors_LHB$Outcome)) +  facet_wrap(~cluster) 
names(abc)
abc_wide <- abc %>%  dplyr::select(genes, Outcome,cluster, value) %>% pivot_wider(names_from = Outcome, values_from = value)
abc_wide %<>% mutate(diff = Good - Bad) %>% 
  mutate(up_down = if_else(diff < 0, "down", "up"))

abc_wide %>% arrange(desc(diff)) %>% group_by(cluster) %>% count() %>% arrange(desc(n))
abc_wide %>% arrange(desc(diff)) %>% group_by(cluster,up_down) %>% count() %>% filter(n > 25) %>% 
  ggplot(aes(up_down, n)) +  geom_bar(stat = "identity", aes(fill = up_down)) +
  facet_wrap(~cluster)



abc_wide %>% abc_wide %>% Good
   gather(key = "Outcome", value = "Expression" , c(Bad, Good)) %>% 
  janitor::tabyl(up_down,Outcome)


abc_wide %>%
  gather(key = "Outcome", value = "Expression" , c(Bad, Good)) %>%  
  ggplot(aes(Outcome, Expression)) +
  geom_point(aes(color = up_down)) + 
   geom_line(aes(group = genes, color = up_down),
              size = 0.1, alpha = 0.5)+
       # scale_color_manual(values = c(colors_LHB$Outcome)) +  
  facet_wrap(~cluster) 

# Use the data yourself for custom figures
ggplot(clust.tab,
       aes(Vac_cycle, value, color = Outcome)) +
  #geom_boxplot(outlier.shape = NA) +
  #geom_point(position = position_jitterdodge(dodge.width = 0.9), size =0.1) +
  geom_line(data = clust.tab %>% filter(Outcome == "Good"),
            aes(group = genes),
             color = "red", size = 0.1, alpha = 0.5) +  
  geom_line(data = clust.tab %>% filter(Outcome == "Bad"),   aes(group = genes),
            color = "grey45", size = 0.1, alpha = 0.5) +  
  scale_color_manual(values = c(colors_LHB$Outcome)) +
  # geom_line(stat = "summary", fun = "median", size = 1.5,
  #           aes(group = Outcome)) +
  scale_color_manual(values = c(colors_LHB$Outcome)) +
  theme_bw() +
  geom_smooth(aes(group=Outcome), method = "auto") + 
  facet_wrap(~cluster) 
```


