---
title: "Autoantidoby manuscript (Baseline analyses)"
output: html_document
date: "2024-09-20"
editor_options: 
  chunk_output_type: console
---

- in this script I will apply limma model to determine which auto-abs are enrichned in Melanoma vs HD samples
at baseline.
- I will perform differential gene expression
- visualizations: volcano plots, counts 

# Paths/libraries/source code
```{r}
# ---- Project setup: paths, packages, colours/themes ----------------------
# Replaces the original machine-specific absolute paths and the two lost
# source files (autoantibody_libraries.R, LHB_color_tools.R). setup.R defines
# file_path, data_path, results_path and loads all packages + figure themes.
if (file.exists("setup.R")) source("setup.R") else source(file.path("R", "setup.R"))
```

# read PHIPseq_list
```{r}
PHIPseq_list <-  readRDS(paste0(data_path, "PHIPseq_list.Rda"))
```

### select subjects
```{r}
PHIPseq_list$df_Phip_metadata$Vac_cycle %>% unique()
d0_pts_HDs <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_0") %>% pull(strings)
d0_pts <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_0" & Outcome != "HD") %>% pull(strings)
d43_pts <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_43") %>% pull(strings)
d89_pts <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_89_101") %>% pull(strings)
d43_pts_Good <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_43" & Outcome == "Good") %>% pull(strings)
d43_pts_Bad <- PHIPseq_list$df_Phip_metadata %>% filter(Vac_cycle == "day_43" & Outcome == "Bad") %>% pull(strings)
```

# --------------
# limma model 

# define matrix of interest
```{r}
# Define a function to subset and format matrix and metadata
prepare_phip_data <- function(data_name, phip_list, pts) {
  # Subset the matrix based on the input dataset name and patients of interest
  matrix_phip <- phip_list[[data_name]] %>% 
    dplyr::select(all_of(intersect(pts, colnames(phip_list[[data_name]]))))
  
  # Subset the metadata to match the columns of the matrix
  meta_phip <- phip_list$df_Phip_metadata %>% 
    filter(strings %in% colnames(matrix_phip))
  
  # Set the row names of the metadata to match the column names of the matrix
  rownames(meta_phip) <- meta_phip$strings
  
  # Return the matrix and metadata as a list
  list(matrix = matrix_phip, metadata = meta_phip)
}

# use of the function and CHOOSE dtaaset
phip_data <- prepare_phip_data("Phip_seq_R3_rpk", PHIPseq_list, c(d0_pts,d43_pts,d89_pts))

# Accessing the resulting matrix and metadata
matrix_phip <- phip_data$matrix
meta_phip <- phip_data$metadata
dim(meta_phip)
####################
# Check if column names of matrix_phip match row names of meta_phip
if (all(colnames(matrix_phip) == rownames(meta_phip))) {
  print("The column names of matrix_phip match the row names of meta_phip.")
} else if (setequal(colnames(matrix_phip), rownames(meta_phip))) {
  print("The column names are the same but in different order. Reordering meta_phip...")
  meta_phip <- meta_phip[colnames(matrix_phip), ]
  print("meta_phip has been reordered to match column names of matrix_phip.")
} else {
  print("The column names do not match the row names of meta_phip. No action taken.")
}

####################
meta_phip$Outcome <- factor(meta_phip$Outcome, levels=c("Good","Bad"))
meta_phip$Cancer <- factor(meta_phip$Cancer, levels=c("Melanoma","HD"))
meta_phip$Vac_cycle <- factor(meta_phip$Vac_cycle, levels = c("day_0","day_43","day_89_101"))
####################

meta_phip %>% dplyr::select(Outcome,Vac_cycle)
meta_phip$Out_cycle <- factor(paste(meta_phip$Outcome,meta_phip$Vac_cycle,sep="_")) 
```



```{r}
# Outcome vs vaccine cycle design  
design <- model.matrix(~0 + Out_cycle, data = meta_phip)
colnames(design) <- levels(meta_phip$Out_cycle)

contrast_matrix <- makeContrasts(
  Bad_day_43_vs_Bad_day_0 = Bad_day_43 - Bad_day_0,
  Good_day_43_vs_Good_day_0 = Good_day_43 - Good_day_0,
  levels = design
)

# Assuming `matrix_phip` is your data matrix with genes/proteins as rows and samples as columns
# log transformation of the data
log_rpk <- log2(matrix_phip + 0.01)  # Adding 0.01 to avoid log(0)
# Fit the linear model to the log-transformed data
vfit <- lmFit(log_rpk, design)
# Apply the contrast matrix
vfit <- contrasts.fit(vfit, contrast_matrix)
# Perform empirical Bayes smoothing to the standard errors
efit <- eBayes(vfit)
# Extract the top table with all results

top_tables_list <- list()

# Loop through each contrast and extract top tables
for (contrast_name in colnames(contrast_matrix)) {
  # Extract top table for the current contrast
  toptab <- topTable(efit, coef = contrast_name, sort.by = "P", n = Inf)
  
  # Store in the list with the contrast name as the key
  top_tables_list[[contrast_name]] <- toptab
  
  # Print or view the top table for the current contrast
  cat("\nTop table for contrast:", contrast_name, "\n")
  print(head(toptab))  # Displaying the first few rows of the top table
}


top_tables_list <- lapply(top_tables_list, function(x) {
  x$Gene <- rownames(x)
  return(x)
})
head(top_tables_list[[1]])
top_tables_list %>% names()

# Create a new list to store the filtered data frames
top_tables_list_filtered <- list()
# Loop through each data frame in the top_tables_list
for (tab_name in names(top_tables_list)) {
  # Access the data frame by name
  tab <- top_tables_list[[tab_name]]
  
  tab <- tab %>% mutate(
  background_gene = if_else(Gene %in% background_genes, "Background", "Not Background")
  )
  # Apply filtering: arrange by logFC and filter based on P.Value and logFC
  filtered_tab <- tab %>% 
    arrange(logFC) %>% 
    filter(P.Value <= 0.05 & abs(logFC) > 1)
  
  # Store the filtered data frame in the new list with the same name
  top_tables_list_filtered[[tab_name]] <- filtered_tab
}

lapply(top_tables_list_filtered, function(x) x %>% janitor::tabyl(background_gene))
lapply(top_tables_list_filtered, nrow)
fin_top_tables_list <- lapply(top_tables_list_filtered, function(x) x %>% filter(background_gene != "Background"))

fin_top_tables_list[[1]]  %>% filter(grepl("TYR|MAGE|MLAN", Gene)) # to check if any of the vaccine antigen peptides may be

genes_change <- unique(c(fin_top_tables_list[[1]]$Gene,fin_top_tables_list[[2]]$Gene))
intersect(fin_top_tables_list[[2]]$Gene,genes_R2_R3)
```

```{r}
limma_toptab_R3 <- limma_toptab
limma_toptab_R2 <- limma_toptab

genes_R2_R3 <- intersect(limma_toptab_R2$Gene,limma_toptab_R3$Gene)
genes_R2_R3 <- genes_R2_R3[!grepl("^LOC", genes_R2_R3)]

genes_R2_R3 %>% saveRDS("genes_R2_R3.Rda")
```

### Figures 1 (volcano)
- https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
```{r}
a <- 1
head(fin_top_tables_list)
# Gene expression table
fin.volcano.tab <- data.frame(fin_top_tables_list[[a]]) %>% arrange(desc(logFC)) #%>% filter(background_gene == "Not Background")
rownames(fin.volcano.tab) <- fin.volcano.tab$Gene
# up in Melanoma
top_sig_genes_up <- fin.volcano.tab %>% filter(logFC > 0 ) %>% 
  #filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)
# up in HD
top_sig_genes_down <- fin.volcano.tab %>% filter(logFC < 0 ) %>% 
  # filter(Gene %in% genes_R2_R3) %>% 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% 
  head(10) %>% 
  pull(Gene)


# Volcano Plot
###########
EnhancedVolcano::EnhancedVolcano(top_tables_list[[a]],
    lab = rownames(top_tables_list[[a]]),
    title = "Baseline comparison",
    subtitle = NULL,
    caption = NULL,
    x = 'logFC', maxoverlapsConnectors = Inf,
    y = 'P.Value',
    selectLab = c(top_sig_genes_up,top_sig_genes_down),
    xlab = bquote(~Log[2]~ 'fold change'),
    ylim = c(0, 5.5),
    pCutoff = 0.05,
    FCcutoff = 0.5,
    pointSize = 3.0,
    labSize = 4.0,
    labCol = 'black',
    labFace = 'plain',
    boxedLabels = TRUE,
    parseLabels = TRUE,
    col = c('black','grey', 'purple', 'red3'),  # Adjust colors if necessary
    colAlpha = 4/5,
    legendLabels = c('Not significant','LogFC > 0.5', 'p-value <= 0.05', 'p-value <= 0.05 & LogFC > 0.5'),
    legendPosition = 'right',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'black') +
    annotate("text", x = c(-3, 3), y = 5.2, label = c("Bad Outcome","Good Outcome"), size = 6)
```

### Figures 1 (counts)
```{r}
R3_rpk_long_wo_HDs <- PHIPseq_list$R3_rpk_long
R3_rpk_long_wo_HDs %<>% filter(Sample %in% c(d0_pts,d43_pts_HDs))
R3_rpk_long_wo_HDs %<>% filter(Outcome != "HD")
R3_rpk_long_wo_HDs$Outcome <- factor(R3_rpk_long_wo_HDs$Outcome, levels=c("Good","Bad"))
R3_rpk_long_wo_HDs$Out_cycle <- factor(paste(R3_rpk_long_wo_HDs$Outcome,R3_rpk_long_wo_HDs$Vac_cycle,sep="_")) 
head(R3_rpk_long_wo_HDs)

top_sig_genes_up
comparisons <- list(c("Bad_day_0", "Bad_day_43"),c("Good_day_0" ,"Good_day_43"))  
gene_int <- c("MR1")
R3_rpk_long_wo_HDs %>%  filter(Expression > 0) %>% 
  #filter(Outcome =="Good") %>% 
  #filter(grepl("pt_", Sample)) %>% 
  dplyr::filter(Gene == gene_int) %>% 
  ggplot(aes(x = Out_cycle, y = log2(Expression+0.01))) +
  geom_boxplot() +
    geom_point(aes(col = Response)) +
        scale_color_manual(values = c(colors_LHB$Response))+
  theme(aspect.ratio = 1) + theme_bw()+
  ggtitle(paste0("Expression of ", gene_int, " gene")) +
  stat_compare_means(
      method = "wilcox.test", 
      comparisons = comparisons,      # List of comparisons to make
      aes(label = sprintf("p = %.2f", as.numeric(..p.format..))),  
      label.y = max_y + 0.5,          
      size = 4.5,                       
      tip.length = 0.02,               
       vjust = 0.05                    
    ) 
  
PHIPseq_list$R3_rpk_long %>% 
  filter(Outcome !="HD") %>% 
  #filter(grepl("pt_", Sample)) %>% 
  dplyr::filter(Gene == gene_int) %>% 
 # left_join(df_Phip_metadata, by = c("Sample" = "strings")) %>% 
  ggplot(aes(x = Vac_cycle, y = log(Expression))) +
  geom_boxplot() +
  geom_line(aes(group = Subject_ID), color = "Grey25", alpha = 0.3) +
  geom_point(aes(col = Response)) +
        scale_color_manual(values = c(colors_LHB$Response))+
  theme(aspect.ratio = 1) + theme_bw()+
  ggtitle(paste0("Expression of ", gene_int, " gene")) +
  facet_wrap(~Outcome)
```

### Multi Box Plots (counts)
```{r}
library(patchwork)  # For combining plots into a grid
jitter_box_1 <- position_jitter(width = 0.2, height = 0)
# Define the genes of interest
gene_int <- c(top_sig_genes_down)

# Define comparisons to be made for the stat_compare_means
comparisons <- list(c("Bad_day_0", "Bad_day_43"),c("Good_day_0" ,"Good_day_43"))  

plot_list <- list()
# Loop over each gene in the gene_int list
for (gene in gene_int) {
  
  # Filter the data for the current gene
  gene_data <- R3_rpk_long_wo_HDs %>%
    dplyr::filter(Gene == gene)
  gene_data$Out_cycle <- factor(gene_data$Out_cycle, levels=c("Bad_day_0", "Bad_day_43","Good_day_0" ,"Good_day_43"))
  # Calculate the maximum y value for the current gene
  max_y <- max(log2(gene_data$Expression+0.01), na.rm = TRUE)
  
  # Create the plot
  p <- gene_data %>%
    ggplot(aes(x = Out_cycle, y = log2(Expression+0.01))) +
    geom_boxplot(outlier.shape = NA) +
    #geom_point(aes(col = Response)) +
     geom_jitter(position = jitter_box_1, aes(col = Response),
                 size = 2, alpha = 0.5) +
    scale_color_manual(values = c(colors_LHB$Response)) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
    
    theme_prism() +
      labs(title = "Baseline PHIP-seq expression",
           # x = "",
           y = "PHIP-seq log(rpk)") +
    theme(
      aspect.ratio = 2/1.7,
      legend.position = "none",
      plot.title = element_text(size = 15),
      axis.title.x = element_blank(),  # Remove x-axis label
      axis.title.y = element_blank()   # Remove y-axis label
    ) +
    stat_compare_means(
      method = "wilcox.test", 
      comparisons = comparisons,      # List of comparisons to make
      aes(label = sprintf("p = %.2f", as.numeric(..p.format..))),  
      label.y = max_y + 0.5,          
      size = 4.5,                       
      tip.length = 0.02,               
       vjust = 0.05                    
    ) +
    ggtitle(paste0(gene))

  plot_list[[gene]] <- p   # Add the plot to the list
}

# Combine all plots into a grid using patchwork
combined_plot <- wrap_plots(plot_list, ncol = 5) 
print(combined_plot)
# combined_plot + plot_layout(guides = "collect") 
#   theme(legend.position = "right")# & labs(x = "Serveur", y = "Distance Parcourue")
```

### Background Expression Statistics 
```{r}
log2_Phip_R3_rpk <- matrix_phip
df_test <- data.frame(Row_sums =rowSums(log2_Phip_R3_rpk))
head(df_test)
background_threshold <- quantile(df_test$Row_sums, 0.05)
df_test <- df_test %>% 
  mutate(Background = ifelse(Row_sums <= background_threshold, "Yes", "No"))
df_test %>% janitor::tabyl(Background)

background_genes <- df_test %>% filter(Background == "Yes") %>% pull(Gene)

ggplot(df_test, aes(x = log2(Row_sums))) + 
  geom_histogram(bins = 250, aes(fill = Background)) +
   scale_fill_manual(values = c("grey40","black"))+
    #geom_vline(xintercept = log2(background_threshold), linetype = "dashed", color = "red", size = 1) +  # Vertical line
  theme_prism() +
  ggtitle("Distribution of Phip-seq Expression Values") +
    labs(title = "Distribution of PHip-seq Expression Values",  
       x = "Log2(rpk)",                         
       y = "count") +        
  theme(aspect.ratio = 1/1)  +
  theme_prism()   

names(Phip_R3_rpk)
head(df_test)

df_test$Healthy <- rowSums(Phip_R3_rpk[, grep("^HD[1-9]|HC[0-9]+", colnames(Phip_R3_rpk))])
df_test$Melanoma_Pts <- rowSums(Phip_R3_rpk[, grep("^pt_", colnames(Phip_R3_rpk))])
df_test$Gene <- rownames(df_test)

#genes_R2_R3 # genes that overlap in LImma analysis btween R2 and R3

limma_toptab_genes <- limma_toptab %>%  # genes from R3 limma analysis 
  filter(!grepl("LOC", Gene)) %>% 
  arrange(P.Value) %>% pull(Gene)

df_test %<>% mutate(
  Significant_gene = if_else(Gene %in% limma_toptab_genes, "Significant", NA)
)
df_test %>% janitor::tabyl(Background,Significant_gene)

df_test %<>% left_join(limma_toptab, by = "Gene")
# Define a vector of genes you want to label
genes_to_label <- c(df_test$Gene)  # Replace with your genes of interest
genes_to_label
# Plot with selected gene labe

df_test %>% filter(Background == "No") %>% filter(Cancer == "Healthy") %>% filter(Significant_gene == "Significant")
df_test %>% filter(Background == "No") %>% filter(Cancer == "Cancer") %>% filter(Significant_gene == "Significant")

head(df_test)
df_test$Cancer %>% unique()
ggplot(df_test, aes(x = log2(Healthy), y = log2(Melanoma_Pts))) +
   geom_point(data = df_test %>% filter(Background == "Yes") %>% filter(is.na(Cancer)),
             color = "black", size = 0.5) +  # First set of points for Healthy
  geom_point(data = df_test %>% filter(Background == "No") %>% filter(is.na(Cancer)), 
             color = "Grey80", size = 0.5) +  # First set of points for Healthy
  geom_point(data = df_test %>% filter(Background == "No") %>% filter(Cancer == "Healthy") %>% 
               filter(Significant_gene == "Significant"),
              color = "#88B8CA", size = 2) +  # First set of points for Healthy
  geom_point(data = df_test %>% filter(Background == "No") %>% filter(Cancer == "Cancer") %>% 
               filter(Significant_gene == "Significant"),
              color = "#FFB6C1", size = 2) +  # Second set of points for Melanoma
  # geom_text(aes(label = ifelse(Gene %in% c(top_sig_genes_up, top_sig_genes_down), Gene, NA)),  # Label significant genes
  #           vjust = 1.5, hjust = 1, check_overlap = TRUE) +
    #geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # 45-degree line
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black", size = 0.5) +  # Correlation line
  #   geom_text_repel(
  #   aes(label = ifelse(Gene %in% c(top_sig_genes_up), Gene, NA)),color = "red4",  # Label significant genes
  #   #vjust = 1.5, hjust = -1,
  #   box.padding = 0.25,     # Padding around labels
  #   point.padding = 0.1,   # Padding around points
  #   segment.color = 'black', # Color of the line segment
  #   max.overlaps = Inf,    # Allow maximum number of overlaps before truncating
  #   na.rm = TRUE           # Remove NA labels
  # ) +
  #     geom_text_repel(
  #   aes(label = ifelse(Gene %in% c(top_sig_genes_down), Gene, NA)),color = "navy",  # Label significant genes
  #   #vjust = 1.5, hjust = -1,
  #   box.padding = 0.25,     # Padding around labels
  #   point.padding = 0.1,   # Padding around points
  #   segment.color = 'black', # Color of the line segment
  #   max.overlaps = Inf,    # Allow maximum number of overlaps before truncating
  #   na.rm = TRUE           # Remove NA labels
  # ) +
  labs(title = "Melanoma Patients vs Healthy Controls",  
       x = "HD Log2(rpk)",                         
       y = "Melanoma Log2(rpk)") +        
  theme(aspect.ratio = 1/1)  +
  theme_prism()                                        
```

```{r}
# selecting good and bad patients at baseline
matrix_phip <- dfa %>% select(all_of(d0_pts_HDs)) %>% 
  select(matches("pt_")) #%>% 
  select(matches("day_0|day_89_101")) 
meta_phip <- PHIPseq_list$df_Phip_metadata %>% filter(strings %in% c(d0_pts_HDs)) %>% 
    filter(grepl("pt_", strings)) #%>% 
  filter(Vac_cycle %in% c("day_0","day_89_101"))
####################
matrix_phip <- dfa %>% select(all_of(d0_pts_HDs)) %>% 
  select(matches("pt_")) %>% 
  select(matches("day_0|day_89_101")) 

meta_phip <- PHIPseq_list$df_Phip_metadata %>% filter(strings %in% c(bad_outcomw_pts)) %>% 
    filter(grepl("pt_", strings)) %>% 
  filter(Vac_cycle %in% c("day_0","day_89_101"))

```
