---
title: "Untitled"
output: html_document
date: "2024-09-19"
editor_options: 
  chunk_output_type: console
---

---
  title: "05_find_tissue_specific_genes_human_protein_atlas"
author: "Ming Tang"
date: "6/15/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---
  
  http://www.bioconductor.org/packages/devel/bioc/vignettes/HPAanalyze/inst/doc/b_HPAanalyze_indepth.html


```{r}
if (file.exists("setup.R")) source("setup.R") else source(file.path("R", "setup.R"))
library(HPAanalyze)
if (file.exists("gene_tissue_classification_source.R")) {
  source("gene_tissue_classification_source.R")
} else source(file.path("R", "gene_tissue_classification_source.R"))
```

# Download HPA datasets
```{r}
downloadedData$normal_tissue %>%
  janitor::tabyl(tissue)

# TISSUE SITE
##############################################################################################################################
# Split the 'Tissues' column into separate rows
# tissues_tbl <- tissues_tbl %>%
#   separate_rows(Tissues, sep = ", ") %>%
#   arrange(Category)

downloadedData$normal_tissue$level
tissue.tab <- downloadedData$normal_tissue %>%   filter(level %in% c("High","Mid","Low")) %>% dplyr::select(Gene = gene, tissue)

tissue.tab %<>%   mutate(tissue_combined = str_replace_all(tissue, c("stomach \\d+" = "stomach", 
                                                     "skin \\d+" = "skin",
                                                     "endometrium \\d+" = "endometrium",
                                                     "soft tissue \\d+" = "soft tissue"))) %>% dplyr::select(-tissue) %>% distinct() %>% 
  dplyr::rename("tissue" = "tissue_combined")
dim(tissue.tab)  
tissue.tab %<>% left_join(anatomical_categories, by = c("tissue" = "Term"))

tissue.tab[, names(tissue.tab) %in% c("Gene","Category")] %>% distinct()
tissue.tab %>% distinct(Category)

# CELLUAR LOCALIZATION
##############################################################################################################################
cell.location.tbl <- downloadedData$subcellular_location %>%
  dplyr::select(Gene = gene, main_location) %>% distinct()

cell.location.tbl %<>% 
  mutate(first_location = sub(";.*", "", main_location)) %>% 
  mutate(first_location = if_else(is.na(first_location), "Mitochondria", first_location)) 

cell.location.tbl %>% pull(first_location) %>% unique()

cell.location.tbl %<>% left_join(cell_structure_categories, by = c("first_location" = "Term")) %>% 
  dplyr::select(-main_location, cellular_location = first_location)
```

### $ CLUST.combined.df
```{r}
combined_df

CLUST.genes.to.tissue.map <- c(CLUST_6_filters,
CLUST_2_filters,
CLUST_9_filters,
CLUST_10_filters,
CLUST_8_filters,
CLUST_5_filters,
CLUST_11_filters,
CLUST_7_filters,
CLUST_3_filters) %>% unique()

CLUST.combined.df <- combined_df %>% filter(Description %in% CLUST.genes.to.tissue.map) 
```

```{r}
###################
CLUST.genes <- CLUST.combined.df %>%
  pull(geneID_split) %>%
  as.character() %>%
  strsplit(" ") %>%    # Split by space
  unlist() %>%  unique()

CLUST.genes %>% length() # 1591 genes 
###################
# clust.tab is original DF from getting DEGPatterns
CLUST.tab <- clust.tab %>% filter(genes %in% CLUST.genes)

HPA_tab <- cell.location.tbl
CLUST.tab.tissue <- CLUST.tab %>% 
  #filter(Gene %in% c(limma_toptab %>% pull(Gene))) %>% 
  left_join(HPA_tab, by = c("genes" = "Gene")) %>% 
  mutate(Category = if_else(is.na(Category), "Not mapped", Category),
         cellular_location = if_else(is.na(cellular_location), "Not mapped", cellular_location))

CLUST.tab.tissue$cellular_location %>% unique()
abc <- CLUST.tab.tissue %>% dplyr::select(-c(Vac_cycle,Outcome,Outcome_cycle,cellular_location,value)) %>% distinct() %>% 
  #filter(grepl("Integument",Category)) %>% 
  #filter(cluster == "1") %>% 
  group_by(genes,cluster) %>% 
  summarize(Category = paste(unique(Category), collapse = ";"), .groups = 'drop') 
  #column_to_rownames("genes") #%>% as.matrix()
head(abc)

aa_meta <- abc %>% pivot_wider(
    names_from = genes, 
    values_from = "Category")

abc$Category %>% unique()
```


# Immmune_annotation
- from REAP data
```{r}
Immmune_annotation <- read_csv(paste0(data_path, "/library_annotation.csv"))
Immmune_annotation %<>% dplyr::select(gene_name, annotation_group:reactome_group) %>% distinct()
Immmune_annotation %<>% filter(!is.na(gene_name))
Immmune_annotation %<>% 
  distinct(gene_name, annotation_group, molfct_group,reactome_group) %>% 
  group_by(gene_name) %>% 
  slice(1) %>% 
  ungroup()
```

- extra code
```{r}
############################
HPA_tab_reform <- HPA_tab %>% 
  group_by(gene) %>%
  summarise(tissues = paste(unique(Category), collapse = ", ")) %>%
  ungroup()

HPA_tab_wide <- HPA_tab %>%
  group_by(Gene) %>%
  mutate(row = row_number()) %>%  # Create a row identifier
  pivot_wider(names_from = row, values_from = Category, names_prefix = "tissue_") %>%
  ungroup()
HPA_tab_wide
##########################################################################################
```

```{r}
tissue_fin_toptab <- limma_toptab %>% arrange(logFC) %>%  filter(P.Value <= 0.05 & abs(logFC) > 0.5)
tissue_fin_toptab %<>%  mutate(Cancer = if_else(logFC > 0, "Cancer", "Healthy"))
tissue_fin_toptab

# CHANGE ME 
HPA_tab <- tissue.tab
# dim(tissue_fin_toptab)
# tissue_toptab <- tissue_fin_toptab %>% left_join(HPA_tab, by = c("Gene"="Gene"))
# tissue_toptab %>% names()
# tissue_toptab %<>% 
#   mutate(Category = if_else(is.na(Category), "Not mapped", Category),
#          tissue = if_else(is.na(tissue), "Not mapped", tissue))
# 
# wide_tissue_toptab <- tissue_toptab %>% dplyr::select(t,Gene,Cancer,tissue) %>% 
#     pivot_wider(names_from = tissue , values_from = tissue, values_fill = NA)
# 
# wide_tissue_mat <- wide_tissue_toptab %>% column_to_rownames("Gene") %>% dplyr::select(-c(t,Cancer)) %>% as.matrix()
```



```{r}
Phip_seq_R3_a$Gene <-  rownames(Phip_seq_R3_a)

Phip_seq_R3_a %<>% 
  filter(Gene %in% c(limma_toptab %>% pull(Gene))) %>% 
  left_join(HPA_tab, by = "Gene") %>% 
  mutate(Category = if_else(is.na(Category), "Not mapped", Category),
         tissue = if_else(is.na(tissue), "Not mapped", tissue))


limma_toptab %<>% mutate(Cancer = if_else(logFC > 0, "Cancer", "Healthy"))
top_100_genes_R2_R3 <- limma_toptab %>% arrange(P.Value) %>% 
  filter(Gene %in% limma_genes) %>% 
  filter(!grepl("LOC",Gene)) %>% 
  filter(logFC > 0.5) %>% head(100) %>% pull(Gene)

long_dff <- Phip_seq_R3_a %>% gather(key = "Sample", value = "Expression", 
                               names(Phip_seq_R3_a[, !names(Phip_seq_R3_a) %in% "Gene"]))

long_dff <- long_dff %>%  
  filter(Sample %in% colnames(matrix_phip) & Gene %in% c(top_100_genes_R2_R3))


long_dff %<>% left_join(filt_limma_toptab %>% dplyr::select(Gene,Cancer), by = "Gene")

long_dff %<>% left_join(HPA_tab, by = "Gene")

long_dff %<>% mutate(Category = if_else(is.na(Category), "Not mapped", Category),
         tissue = if_else(is.na(tissue), "Not mapped", tissue))
long_dff$Category %>% unique()

long_dff %<>% mutate(Expression = as.numeric(Expression))

# Step 1: Combine multiple category values into a single column separated by ";"
df_combined <- long_dff %>% filter(Category != "Not mapped") %>% 
   mutate(Category = if_else(abs(Expression) < 2, "", Category)) %>% 
  group_by(Gene, Sample) %>% 
  summarize(Category = paste(unique(Category), collapse = ";"), .groups = 'drop')

# Step 2: Pivot the table to a wider format
df_wide <- df_combined %>%
  pivot_wider(
    names_from = Sample,     # Sample column as the new column names
    values_from = Category   # Values filled from the combined Category column
  )

matrix_JA <- df_wide %>% column_to_rownames("Gene") %>% as.matrix()
dim(matrix_JA)


matrix_JA <- long_dff %>%   filter(Category != "Not mapped") %>% 
  group_by(Gene, Sample) %>% 
  summarize(Category = paste(unique(Category), collapse = ";"), .groups = 'drop') %>% column_to_rownames("Gene") %>% as.matrix()


meta_JA <- long_dff %>% mutate(Category = if_else(abs(Expression) < 1, "", Category)) %>% 
  dplyr::select(Gene,Sample,Category ) %>% 
  filter(Category != "Not mapped") %>% 
  pivot_wider(
    names_from = Sample,     # Sample column as the new column names
    values_from = Category     # Values filled from the tissue column
  )

meta_JA <- df_wide %>% left_join(HPA_tab, by = "Gene") %>% 
   mutate(Category = if_else(is.na(Category), "Not mapped", Category),
         tissue = if_else(is.na(tissue), "Not mapped", tissue))
```

# my attempt oncoprint
### 1 cellular structures
```{r}
# Extend the color vector with missing categories and assign colors
col <- c(
  "Not mapped" = "#d3d3d3",            # Light gray for not mapped
  "Nucleoplasm" = "#1f78b4",           # Blue
  "Mitochondria" = "#33a02c",          # Green
  "Actin filaments" = "#e31a1c",       # Red
  "Vesicles" = "#ff7f00",              # Orange
  "Plasma membrane" = "#6a3d9a",       # Purple
  "Centrosome" = "#b15928",            # Brown
  "Golgi apparatus" = "#a6cee3",       # Light blue
  "Nuclear bodies" = "#fb9a99",        # Pink
  "Cytosol" = "#b2df8a",               # Light green
  "Nucleoli" = "#fdbf6f",              # Light orange
  "Focal adhesion sites" = "#cab2d6",  # Light purple
  "Nuclear speckles" = "#ffff99",      # Yellow
  "Nucleoli fibrillar center" = "#1f78b4",  # Blue (same as Nucleoplasm)
  "Microtubules" = "#33a02c",          # Green (same as Mitochondria)
  "Intermediate filaments" = "#e31a1c",# Red (same as Actin filaments)
  "Centriolar satellite" = "#6a3d9a",  # Purple (same as Plasma membrane)
  "Endoplasmic reticulum" = "#ff7f00", # Orange (same as Vesicles)
  "Cell Junctions" = "#b15928",        # Brown (same as Centrosome)
  "Cytoplasmic bodies" = "#a6cee3",    # Light blue (same as Golgi apparatus)
  "Nuclear membrane" = "#fb9a99",      # Pink (same as Nuclear bodies)
  "Lipid droplets" = "#b2df8a",        # Light green (same as Cytosol)
  "Midbody" = "#fdbf6f",               # Light orange (same as Nucleoli),
  "Aggresome" = "#ff6347",             # Tomato red for Aggresome
  "Endosomes" = "#4682b4",             # Steel blue for Endosomes
  "Mitotic chromosome" = "#ff4500",    # Orange red for Mitotic chromosome
  "Nucleoli rim" = "#deb887",          # Burlywood for Nucleoli rim
  "Cytokinetic bridge" = "#8b4513",    # Saddle brown for Cytokinetic bridge
  "Peroxisomes" = "#7fffd4"            # Aquamarine for Peroxisomes
)

# Extend the alter_fun list with missing categories
alter_fun <- list(
  background = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = "#CCCCCC", col = NA))
  },
  
  # Existing alterations
  `Not mapped` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Not mapped"], col = NA)) },
  Nucleoplasm = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nucleoplasm"], col = NA)) },
  Mitochondria = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Mitochondria"], col = NA)) },
  `Actin filaments` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h*0.33, gp = gpar(fill = col["Actin filaments"], col = NA)) },
  Vesicles = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Vesicles"], col = NA)) },
  `Plasma membrane` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Plasma membrane"], col = NA)) },
  Centrosome = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Centrosome"], col = NA)) },
  `Golgi apparatus` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Golgi apparatus"], col = NA)) },
  `Nuclear bodies` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nuclear bodies"], col = NA)) },
  Cytosol = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cytosol"], col = NA)) },
  Nucleoli = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nucleoli"], col = NA)) },
  `Focal adhesion sites` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Focal adhesion sites"], col = NA)) },
  `Nuclear speckles` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nuclear speckles"], col = NA)) },
  `Nucleoli fibrillar center` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nucleoli fibrillar center"], col = NA)) },
  Microtubules = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Microtubules"], col = NA)) },
  `Intermediate filaments` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h*0.33, gp = gpar(fill = col["Intermediate filaments"], col = NA)) },
  `Centriolar satellite` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Centriolar satellite"], col = NA)) },
  `Endoplasmic reticulum` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Endoplasmic reticulum"], col = NA)) },
  `Cell Junctions` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cell Junctions"], col = NA)) },
  `Cytoplasmic bodies` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cytoplasmic bodies"], col = NA)) },
  `Nuclear membrane` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nuclear membrane"], col = NA)) },
  `Lipid droplets` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Lipid droplets"], col = NA)) },
  Midbody = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Midbody"], col = NA)) },
  
  # Newly added categories
  Aggresome = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Aggresome"], col = NA)) },
  Endosomes = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Endosomes"], col = NA)) },
  `Mitotic chromosome` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Mitotic chromosome"], col = NA)) },
  `Nucleoli rim` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nucleoli rim"], col = NA)) },
  `Cytokinetic bridge` = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cytokinetic bridge"], col = NA)) },
  Peroxisomes = function(x, y, w, h) { grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Peroxisomes"], col = NA)) }
)

# Use this updated alter_fun and col in your ComplexHeatmap function as needed

```

### 2 cellular categories
```{r}
# Define colors for each category
col <- c(
  # "Not mapped" = "#d3d3d3",                            # Light gray
  "Cytoplasm and Associated Structures" = "#1f78b4",  # Blue
  "Nucleus and Nuclear Structures" = "#33a02c",       # Green
  "Organelles" = "#e31a1c",                           # Red
  "Membrane-associated Structures" = "#ff7f00",       # Orange
  "Cytoskeleton" = "#6a3d9a"                          # Purple
)
# Create the alter_fun list for each category
alter_fun <- list(
  # Default background
  background = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = "#CCCCCC", col = NA))
  },
  
  # Not mapped
  `Not mapped` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Not mapped"], col = NA))
  },
  
  # Cytoplasm and Associated Structures
  `Cytoplasm and Associated Structures` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cytoplasm and Associated Structures"], col = NA))
  },
  
  # Nucleus and Nuclear Structures
  `Nucleus and Nuclear Structures` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Nucleus and Nuclear Structures"], col = NA))
  },
  
  # Organelles
  Organelles = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Organelles"], col = NA))
  },
  
  # Membrane-associated Structures
  `Membrane-associated Structures` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Membrane-associated Structures"], col = NA))
  },
  
  # Cytoskeleton
  Cytoskeleton = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), gp = gpar(fill = col["Cytoskeleton"], col = NA))
  }
)

```

### 3 tissue categories
```{r}
# Define colors for each category
col <- c(
  "Not mapped" = "#d3d3d3",                   # Light gray
  "Nervous System - Central" = "#1f78b4",     # Blue
  "Digestive System" = "#33a02c",             # Green
  "Reproductive System - Female" = "#e31a1c", # Red
  "Respiratory System" = "#ff7f00",           # Orange
  "Integumentary System" = "#6a3d9a",         # Purple
  "Reproductive System - Male" = "#b15928",   # Brown
  "Immune and Lymphatic System" = "magenta",  # magenta
  "Endocrine System" = "#fb9a99",             # Pink
  "Urinary System" = "#b2df8a",               # Light green
  "Musculoskeletal System" = "#fdbf6f",       # Light orange
  "Nervous System - Peripheral" = "#cab2d6",   # Light purple
  "NA" = "white"
)

library(grid)

# Create the alter_fun list for each category
alter_fun <- list(
  # Default background function
  background = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = "#CCCCCC", col = NA))
  },
  
  # Not mapped
  `Not mapped` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Not mapped"], col = NA))
  },
  
  # Nervous System - Central
  `Nervous System - Central` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Nervous System - Central"], col = NA))
  },
  
  # Digestive System
  `Digestive System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Digestive System"], col = NA))
  },
  
  # Reproductive System - Female
  `Reproductive System - Female` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Reproductive System - Female"], col = NA))
  },
  
  # Respiratory System
  `Respiratory System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Respiratory System"], col = NA))
  },
  
  # Integumentary System
  `Integumentary System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Integumentary System"], col = NA))
  },
  
  # Reproductive System - Male
  `Reproductive System - Male` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Reproductive System - Male"], col = NA))
  },
  
  # Immune and Lymphatic System
  `Immune and Lymphatic System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Immune and Lymphatic System"], col = NA))
  },
  
  # Endocrine System
  `Endocrine System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Endocrine System"], col = NA))
  },
  
  # Urinary System
  `Urinary System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Urinary System"], col = NA))
  },
  
  # Musculoskeletal System
  `Musculoskeletal System` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Musculoskeletal System"], col = NA))
  },
  
  # Nervous System - Peripheral
  `Nervous System - Peripheral` = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["Nervous System - Peripheral"], col = NA))
  },
    # Nervous System - Peripheral
  "NA" = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = col["NA"], col = NA))
  }
)

```

```{r}
# Create the alter_fun list for each category with offset positions
alter_fun <- list(
  # Default background function
  background = function(x, y, w, h) {
    grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
              gp = gpar(fill = "#CCCCCC", col = NA))
  },
  
  # Not mapped: Top-left corner
  `Not mapped` = function(x, y, w, h) {
    grid.rect(x - w * 0.25, y + h * 0.25, w*0.3, h*0.3, 
              gp = gpar(fill = col["Not mapped"], col = NA))
  },
  
  # Nervous System - Central: Top-right corner
  `Nervous System - Central` = function(x, y, w, h) {
    grid.rect(x + w * 0.25, y + h * 0.25, w*0.3, h*0.3, 
              gp = gpar(fill = col["Nervous System - Central"], col = NA))
  },
  
  # Digestive System: Bottom-left corner
  `Digestive System` = function(x, y, w, h) {
    grid.rect(x - w * 0.25, y - h * 0.25, w*0.3, h*0.3, 
              gp = gpar(fill = col["Digestive System"], col = NA))
  },
  
  # Reproductive System - Female: Bottom-right corner
  `Reproductive System - Female` = function(x, y, w, h) {
    grid.rect(x + w * 0.25, y - h * 0.25, w*0.3, h*0.3, 
              gp = gpar(fill = col["Reproductive System - Female"], col = NA))
  },
  
  # Respiratory System: Center
  `Respiratory System` = function(x, y, w, h) {
    grid.rect(x, y, w*0.3, h*0.3, 
              gp = gpar(fill = col["Respiratory System"], col = NA))
  },
  
  # Integumentary System: Top-center
  `Integumentary System` = function(x, y, w, h) {
    grid.rect(x, y + h * 0.35, w*0.3, h*0.3, 
              gp = gpar(fill = col["Integumentary System"], col = NA))
  },
  
  # Reproductive System - Male: Bottom-center
  `Reproductive System - Male` = function(x, y, w, h) {
    grid.rect(x, y - h * 0.35, w*0.3, h*0.3, 
              gp = gpar(fill = col["Reproductive System - Male"], col = NA))
  },
  
  # Immune and Lymphatic System: Left-center
  `Immune and Lymphatic System` = function(x, y, w, h) {
    grid.rect(x - w * 0.35, y, w*0.3, h*0.3, 
              gp = gpar(fill = col["Immune and Lymphatic System"], col = NA))
  },
  
  # Endocrine System: Right-center
  `Endocrine System` = function(x, y, w, h) {
    grid.rect(x + w * 0.35, y, w*0.3, h*0.3, 
              gp = gpar(fill = col["Endocrine System"], col = NA))
  },
  
  # Urinary System: Top-left inside
  `Urinary System` = function(x, y, w, h) {
    grid.rect(x - w * 0.15, y + h * 0.15, w*0.3, h*0.3, 
              gp = gpar(fill = col["Urinary System"], col = NA))
  },
  
  # Musculoskeletal System: Top-right inside
  `Musculoskeletal System` = function(x, y, w, h) {
    grid.rect(x + w * 0.15, y + h * 0.15, w*0.3, h*0.3, 
              gp = gpar(fill = col["Musculoskeletal System"], col = NA))
  },
  
  # Nervous System - Peripheral: Bottom-left inside
  `Nervous System - Peripheral` = function(x, y, w, h) {
    grid.rect(x - w * 0.15, y - h * 0.15, w*0.3, h*0.3, 
              gp = gpar(fill = col["Nervous System - Peripheral"], col = NA))
  },
    # Nervous System - Peripheral: Bottom-left inside
  "NA" = function(x, y, w, h) {
    grid.rect(x - w * 0.15, y - h * 0.15, w*0.3, h*0.3, 
              gp = gpar(fill = col["NA"], col = NA))
  }
)

```

```{r}
library(ComplexHeatmap)
library(circlize)

aa <- aa_meta %>% column_to_rownames("cluster")
# aa[is.na(aa)] <- ""

oncoPrint(aa,
    alter_fun = alter_fun, col = col,     
    #column_split = mata_d$Cancer,
    #row_split = aa_meta$cluster,
    show_row_names = TRUE, na_col = "white",
            height = unit(100, "mm"),
        width = unit(150, "mm"))
```

# MR PLOT
```{r}
colnames(matrix_JA) == meta_phip$strings
# Create the heatmap with the defined legend parameters

mata_d <- data.frame(strings = colnames(matrix_JA)) %>% left_join(meta_phip, by = "strings")

mel_pts <- mata_d %>% filter(Cancer == "Melanoma") %>% pull(strings)
HD_pts <- mata_d %>% filter(Cancer == "HD") %>% pull(strings)

mata_d$Cancer
colnames(matrix_JA) == mata_d$strings

mel_pts_mat <- matrix_JA[, colnames(matrix_JA) %in% mel_pts]
mel_HD_mat <- matrix_JA[, colnames(matrix_JA) %in% HD_pts]


h1 <- oncoPrint(mel_pts_mat,
    alter_fun = alter_fun, col = col,     
    #column_split = mata_d$Cancer,
    #row_split = meta_JA$Category,
    show_row_names = TRUE,
            height = unit(150, "mm"),
        width = unit(75, "mm"))

h2 <- oncoPrint(mel_HD_mat,
    alter_fun = alter_fun, col = col,     
    #column_split = mata_d$Cancer,
    #row_split = meta_JA$Category,
    show_row_names = TRUE,
            height = unit(150, "mm"),
        width = unit(75, "mm"))

h1+h2
```



```{r}
tabbb %>% add_row("Not mapped")

rbind(
  tabbb,
  data.frame(Gene = "Not mapped", Category = "Not mapped")
)


setdiff(colnames(wide_tissue_mat), tabbb$tissue)

wide_tissue_mat %>% dim()
wide_tissue_toptab %>% dplyr::select(Gene) %>% left_join(tissue.tab %>% dplyr::select(Gene, Category), by = "Gene") 
wide_tissue_toptab
wide_tissue_mat
tabbb <- distinct(tissue.tab[,-1])

# Create a row annotation for the heatmap based on the Category
row_anno <- ComplexHeatmap::rowAnnotation(
  Category = tissues_tbl$Category
)

row_anno_2 <- ComplexHeatmap::rowAnnotation(
  Category = tissues_tbl$Tissues
)

Response_anno = ComplexHeatmap::rowAnnotation(Response = Phip_mx_df$Response,
                        simple_anno_size = unit(c(0.4), "cm"),
                    gap = unit(c(1,1), "mm"), annotation_name_gp= gpar(fontsize = 14),
                     show_annotation_name = F,
                        col = list(Response = c(colors_LHB$Response)))
################

ComplexHeatmap::Heatmap(t(my.df_mat),
        cluster_columns  = T, 
        cluster_rows  = T,
        show_column_names = FALSE,
        show_row_names = FALSE,
        row_gap = unit(1, "mm"),
        column_title = "expression",
        height = unit(100, "mm"),
        width = unit(150, "mm"),
        left_annotation = c(row_anno),
        column_split = my.df$Cancer,
        #row_split = colnames(my.df_mat),
        column_labels = T)
```



# GOOD Plots
```{r}
tissue_fin_toptab <- fin_toptab
tissue_fin_toptab %<>%  mutate(Cancer = if_else(logFC > 0, "Cancer", "Healthy"))
tissue_fin_toptab

# CHANGE ME 
HPA_tab <- cell.location.tbl
dim(tissue_fin_toptab)
tissue_toptab <- tissue_fin_toptab %>% left_join(HPA_tab, by = c("Gene"="Gene"))
tissue_toptab %<>% 
  mutate(Category = if_else(is.na(Category), "Not mapped", Category),
         cellular_location = if_else(is.na(cellular_location), "Not mapped", cellular_location))

tissue_toptab$Category <- factor(tissue_toptab$Category, levels = c("Membrane-associated Structures", 
                                              "Cytoplasm and Associated Structures",
                                              "Cytoskeleton",
                                              "Organelles",
                                              "Nucleus and Nuclear Structures",
                                              "Not mapped"))

tissue_toptab %>%
  mutate(Cancer = if_else(logFC > 0, "Melanoma", "HD")) %>% 
  #filter(!is.na(cellular_location)) %>% 
  group_by(Cancer,cellular_location,Category) %>%
  dplyr::count() %>% filter(n >= 1) %>% 
  ggplot(aes(x = fct_reorder(cellular_location,n), y = n)) +
  geom_bar(stat = "identity") +
  labs(x = "Tissue", y = "Count", title = "Counts of Each Tissue Type") +
  theme_bw() +
  guides(fill = guide_legend(title = "LogFC")) +
  theme(plot.title = element_text(hjust = 0.5, size = 20),
        axis.title = element_text(size = 15)) +
  theme(axis.text.x = element_blank(), #element_text(angle = 45, vjust = 1, hjust = 1),
        axis.ticks.x = element_blank(),  # Remove x-axis ticks
        strip.text.x = element_text(angle = 0,size = 18),  # Set right facet strip to horizontal
         strip.text.y = element_text(angle = 0,size = 12),  # Set right facet strip to horizontal
        strip.background = element_rect(fill = "white", color = "white"),  # White background for facet labels and no border
        panel.background = element_rect(fill = "white"),  # Make panels white
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  facet_grid(Category ~ Cancer, scales = "free_y", space = "free_y")+ coord_flip()
  #facet_grid(Category ~ Cancer, scales = "free_y", space = "free_y") +

# select top genes to plot 
top_genes_plots <- tissue_toptab %>% dplyr::select(Gene,t,P.Value) %>% distinct() %>% 
  arrange(P.Value) #%>% head(150)
tissue_toptab$cellular_location

# Summarize the data to get the average or max 't' value for each gene
gene_order <- tissue_toptab %>%
  filter(Gene %in% top_genes_plots$Gene) %>%
  group_by(Gene) %>%
  summarize(avg_t = mean(t, na.rm = TRUE)) %>%
  arrange(desc(avg_t)) %>%
  pull(Gene)

tissue_order <- tissue_toptab %>%
  filter(!is.na(cellular_location)) %>%
  group_by(cellular_location) %>%
  dplyr::count() %>%
  filter(n > 0) %>%
  
  arrange(desc(n)) %>%
  pull(cellular_location)

# Create the plot with reordered tissues and genes based on counts
tissue_toptab %>%
  filter(Gene %in% top_genes_plots$Gene) %>%
    ggplot(aes(fct_reorder(Gene, match(Gene, gene_order)), cellular_location)) + 
  # ggplot(aes(fct_reorder(cellular_location, match(cellular_location, tissue_order)), Gene)) + 
  geom_tile(aes(fill = t)) +
  scale_fill_gradientn(breaks = c(min(top_genes_plots$t), 0, max(top_genes_plots$t)), colors = c("blue", "grey50", "orange")) +
  labs(x = 'Tissue', y = 'Gene') +
  theme_bw() +
  guides(fill = guide_legend(title = "LogFC")) +
  theme(plot.title = element_text(hjust = 0.5, size = 20),
        axis.title = element_text(size = 15)) +
  theme(axis.text.x = element_blank(), #element_text(angle = 45, vjust = 1, hjust = 1),
        axis.ticks.x = element_blank(),  # Remove x-axis ticks
        strip.text.x = element_text(angle = 0,size = 18),  # Set right facet strip to horizontal
         strip.text.y = element_text(angle = 0,size = 12),  # Set right facet strip to horizontal
        strip.background = element_rect(fill = "white", color = "white"),  # White background for facet labels and no border
        panel.background = element_rect(fill = "white"),  # Make panels white
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  facet_grid(Category ~ Cancer, scales = "free", space = "free")


tissue_toptab_wide <- tissue_toptab %>% dplyr::select(logFC,Gene,Cancer,cellular_location) %>% 
  distinct() %>%  # Ensure unique gene-tissue pairs
  pivot_wider(names_from = cellular_location , values_from = logFC, values_fill = NA)

tissue_toptab$Gene_Cell_location <- paste(tissue_toptab$Gene, tissue_toptab$cellular_location, sep = "_")
#tissue_toptab %<>% filter(tissue_combined %in% top_tissues)

# HEATMAP PATIENT INFO
######################################
# Subset the dataframe using rownames in all_row_names
sub_df <- matrix_phip[rownames(matrix_phip) %in% tissue_toptab$Gene, ]

sub_df_long <- sub_df %>%
  pivot_longer(cols = colnames(sub_df), 
               names_to = "strings",  # Name for the column that stores original column names
               values_to = "expression")  # Name for the column that stores data values

sub_df_long %<>% left_join(PHIPseq_list$df_Phip_metadata, by = "strings")
sub_df_long_meta <- sub_df_long %>% dplyr::select(-expression) %>% distinct() %>% column_to_rownames("strings")

colnames(sub_df) == rownames(sub_df_long_meta)
sub_df_mat <- as.matrix(sub_df)
nrow(sub_df_mat)
sub_df$Gene <- rownames(sub_df)
sub_df %<>% left_join(tissue_toptab, by = "Gene")
sub_df %>% filter(!is.na(Category)) %>% dim()
sub_df %>% distinct(Gene) %>% nrow()

library(circlize)
library(grid)

row_anno <- ComplexHeatmap::rowAnnotation(
  Category = anno_text(
    sub_df$Category,  # Assuming sub_df has the 'Category' column for annotation labels
    rot = 90          # Rotate the labels 90 degrees
  )
)

Response_anno = ComplexHeatmap::columnAnnotation(Response = sub_df_long_meta$Response,
                        simple_anno_size = unit(c(0.4), "cm"),
                    gap = unit(c(1,1), "mm"), annotation_name_gp= gpar(fontsize = 14),
                     show_annotation_name = F,
                        col = list(Response = c(colors_LHB$Response)))

ComplexHeatmap::Heatmap(sub_df_mat,
        cluster_columns  = T, 
        row_title_rot = 0,
        cluster_rows  = T,
        cluster_row_slices = FALSE,
        show_column_names = FALSE,
        show_row_names = FALSE,
        row_gap = unit(1, "mm"),
        column_title = "expression",
        width = unit(75, "mm"),
        height = unit(125, "mm"),
        top_annotation = Response_anno,
        column_split = sub_df_long_meta$Cancer,
        left_annotation = c(row_anno),
        row_split = sub_df$Category,
        column_labels = T)



# HEATMAP TISSUE EXPRESSION PATTERNS
######################################

library(oncoprint)
BiocManager::install("oncoprint")



full_tab <- tissue_toptab %>% dplyr::select(-c(logFC:B)) %>% left_join(sub_df, by = "Gene")

rownames(full_tab) <-  full_tab$Gene_Tissue

heat_mat <- full_tab %>% dplyr::select(matches("HC|HD|pt_")) %>% as.matrix()
dim(heat_mat)
meta_phip %>% dim()


tissue_toptab

my.df <- tissue_toptab_wide %>% as.data.frame() %>% column_to_rownames("Gene")
my.df_mat <- my.df %>%  dplyr::select(-Cancer,-"NA") %>% as.matrix()
my.df_mat[is.na(my.df_mat)] <- 0
keep <- abs(rowSums(my.df_mat)) > 0
my.df_mat <- my.df_mat[keep,]
my.df_mat <- abs(my.df_mat)

sub_df <- my.df[rownames(my.df) %in% rownames(my.df_mat), ]
dim(sub_df)

meta_phip %>% dim()

################
# Create a row annotation for the heatmap based on the Category
row_anno <- ComplexHeatmap::rowAnnotation(
  Category = tissues_tbl$Category
)

row_anno_2 <- ComplexHeatmap::rowAnnotation(
  Category = tissues_tbl$Tissues
)

Response_anno = ComplexHeatmap::rowAnnotation(Response = Phip_mx_df$Response,
                        simple_anno_size = unit(c(0.4), "cm"),
                    gap = unit(c(1,1), "mm"), annotation_name_gp= gpar(fontsize = 14),
                     show_annotation_name = F,
                        col = list(Response = c(colors_LHB$Response)))
################

ComplexHeatmap::Heatmap(t(my.df_mat),
        cluster_columns  = T, 
        cluster_rows  = T,
        show_column_names = FALSE,
        show_row_names = FALSE,
        row_gap = unit(1, "mm"),
        column_title = "expression",
        height = unit(100, "mm"),
        width = unit(150, "mm"),
        left_annotation = c(row_anno),
        column_split = my.df$Cancer,
        #row_split = colnames(my.df_mat),
        column_labels = T)

data.frame(Tissues = colnames(my.df_meta)) %>%  left_join(tissues_tbl, by = "Tissues") 

##########################################################################################





intersect(HPA_tab_reform$gene,filt_toptab_genne_anno$Gene)

tab_x <- filt_toptab_genne_anno %>%
  left_join(HPA_wide %>% 
              dplyr::select(`Gene name`, Tissue,`Cell type`), by = c("Gene" = "Gene name")) %>%
  mutate(tissue_expr = coalesce(Tissue, NA)) %>% distinct() %>% 
  dplyr::select(-Tissue)  # Optionally remove the extra 'Tissue' column if you don't need it
tab_x %>%  filter(is.na(tissue_expr))
tab_x %>% nrow()


tissue_list <- list()
tissue_list <- downloadedData$normal_tissue$tissue %>% unique()
View(data)

hpaVis(downloadedData,
       targetGene = c("GCH1", "PTS", "SPR", "DHFR"),
       targetTissue = c("cerebellum", "cerebral cortex", "hippocampus"),
       targetCancer = c("glioma"))

filt_toptab_genne_anno %>%  mutate(Cancer = if_else(logFC > 0, "Melanoma", "HD"))
hpaVisTissue(downloadedData,
             targetGene= grep("LOC",filt_toptab_genne_anno$Gene, value = T,invert = TRUE),
             targetTissue= tissue_list)
```



# Oncoprint
```{r}
mat = read.table(textConnection(
                "s1,s2,s3
                g1,snv;indel,snv,indel
                g2,,snv;indel,snv
                g3,snv,,indel;snv"), 
                row.names = 1, header = TRUE, sep = ",", stringsAsFactors = FALSE)
mat = as.matrix(mat)
mat
```

```{r}
get_type_fun = function(x) strsplit(x, ";")[[1]]
get_type_fun(mat[1, 1])
```

```{r}
mat[1, 1]
get_type_fun(mat[1, 1])
```

```{r}
library(ComplexHeatmap)
oncoPrint(mat)

col = c(snv = "yellow", indel = "purple")
oncoPrint(mat,
    alter_fun = list(
        snv = function(x, y, w, h) grid.rect(x*0.999, y, w*0.9, h*0.9, 
            gp = gpar(fill = col["snv"], col = "black")),
        
        indel = function(x, y, w, h) grid.rect(x, y, w*0.9, h*0.4, 
            gp = gpar(fill = col["indel"], col = "green"))), 
    col = col,
    alter_fun_is_vectorized = FALSE)
```
# Chapter 2 - Using oncoprint in real life

Requirements of input data 
```{r}
mat = read.table(system.file("extdata", package = "ComplexHeatmap", 
    "tcga_lung_adenocarcinoma_provisional_ras_raf_mek_jnk_signalling.txt"), 
    header = TRUE, stringsAsFactors = FALSE, sep = "\t")
mat[is.na(mat)] = ""
rownames(mat) = mat[, 1]
mat = mat[, -1]
mat=  mat[, -ncol(mat)]
mat = t(as.matrix(mat))

mat[1:10, 1:10]
```
making the alter_fun structure 
```{r}
col = c("HOMDEL" = "blue", "AMP" = "red", "MUT" = "#008000")

alter_fun = list(
    background = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = "#CCCCCC", col = NA))
    },
    # big blue
    HOMDEL = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = col["HOMDEL"], col = NA))
    },
    # big red
    AMP = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = col["AMP"], col = NA))
    },
    # small green
    MUT = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h*0.33, 
            gp = gpar(fill = col["MUT"], col = NA))
    }
)
```


```{r}
column_title = "OncoPrint for TCGA Lung Adenocarcinoma, genes in Ras Raf MEK JNK signalling"

heatmap_legend_param = list(title = "Alternations", at = c("HOMDEL", "AMP", "MUT"), 
        labels = c("Deep deletion", "Amplification", "Mutation"))
```


```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))
    
#You can pass Heatmap() arguments in oncoPrint 
```
Removing empty col and rows
```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param)
```
```{r}
sample_order = scan(paste0(system.file("extdata", package = "ComplexHeatmap"), 
    "/sample_order.txt"), what = "character")

oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    row_order = 1:nrow(mat), column_order = sample_order,
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param)
```
```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    top_annotation = HeatmapAnnotation(
        column_barplot = anno_oncoprint_barplot("MUT", border = TRUE, # only MUT
            height = unit(4, "cm"))
    ),
    right_annotation = rowAnnotation(
        row_barplot = anno_oncoprint_barplot(c("AMP", "HOMDEL"),  # only AMP and HOMDEL
            border = TRUE, height = unit(4, "cm"), 
            axis_param = list(side = "bottom", labels_rot = 90))
    ),
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))
```
oncoprint produce heatmap object, you can add them vertically or horizontally directly
```{r}
ht1 <- oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    top_annotation = HeatmapAnnotation(
        column_barplot = anno_oncoprint_barplot("MUT", border = TRUE, # only MUT
            height = unit(4, "cm"))
    ),
    right_annotation = rowAnnotation(
        row_barplot = anno_oncoprint_barplot(c("AMP", "HOMDEL"),  # only AMP and HOMDEL
            border = TRUE, height = unit(4, "cm"), 
            axis_param = list(side = "bottom", labels_rot = 90))
    ),
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))


ht1 + ht1 
```


```{r}
# oncoPrint(mat,
#     alter_fun = alter_fun, col = col, 
#     column_title = column_title, heatmap_legend_param = heatmap_legend_param)
# 
# oncoPrint(mat,
#     alter_fun = alter_fun, col = col, 
#     column_title = column_title, heatmap_legend_param = heatmap_legend_param,
#     column_title_gp = gpar(fontsize = 9, fontface = "bold"))


```


```{r}
colnames(wide_tissue_mat)
wide_tissue_toptab
wide_tissue_mat
wide_tissue_mat[is.na(wide_tissue_mat)] <- ""

oncoPrint(wide_tissue_mat)
```

```{r}
mat = read.table(system.file("extdata", package = "ComplexHeatmap", 
    "tcga_lung_adenocarcinoma_provisional_ras_raf_mek_jnk_signalling.txt"), 
    header = TRUE, stringsAsFactors = FALSE, sep = "\t")
mat[is.na(mat)] = ""
rownames(mat) = mat[, 1]
mat = mat[, -1]
mat=  mat[, -ncol(mat)]
mat = t(as.matrix(mat))

mat[1:10, 1:10]
```

making the alter_fun structure 
```{r}
col = c("HOMDEL" = "blue", "AMP" = "red", "MUT" = "#008000")

alter_fun = list(
    background = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = "#CCCCCC", col = NA))
    },
    # big blue
    HOMDEL = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = col["HOMDEL"], col = NA))
    },
    # big red
    AMP = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h-unit(2, "pt"), 
            gp = gpar(fill = col["AMP"], col = NA))
    },
    # small green
    MUT = function(x, y, w, h) {
        grid.rect(x, y, w-unit(2, "pt"), h*0.33, 
            gp = gpar(fill = col["MUT"], col = NA))
    }
)
```


```{r}
column_title = "OncoPrint for TCGA Lung Adenocarcinoma, genes in Ras Raf MEK JNK signalling"

heatmap_legend_param = list(title = "Alternations", at = c("HOMDEL", "AMP", "MUT"), 
        labels = c("Deep deletion", "Amplification", "Mutation"))
```


```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    column_title = column_title, heatmap_legend_param = heatmap_legend_param)

oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))
    
#You can pass Heatmap() arguments in oncoPrint 
```
Removing empty col and rows
```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param)
```
```{r}
sample_order = scan(paste0(system.file("extdata", package = "ComplexHeatmap"), 
    "/sample_order.txt"), what = "character")

oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    row_order = 1:nrow(mat), column_order = sample_order,
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param)
```
```{r}
oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    top_annotation = HeatmapAnnotation(
        column_barplot = anno_oncoprint_barplot("MUT", border = TRUE, # only MUT
            height = unit(4, "cm"))
    ),
    right_annotation = rowAnnotation(
        row_barplot = anno_oncoprint_barplot(c("AMP", "HOMDEL"),  # only AMP and HOMDEL
            border = TRUE, height = unit(4, "cm"), 
            axis_param = list(side = "bottom", labels_rot = 90))
    ),
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))
```
oncoprint produce heatmap object, you can add them vertically or horizontally directly
```{r}
ht1 <- oncoPrint(mat,
    alter_fun = alter_fun, col = col, 
    top_annotation = HeatmapAnnotation(
        column_barplot = anno_oncoprint_barplot("MUT", border = TRUE, # only MUT
            height = unit(4, "cm"))
    ),
    right_annotation = rowAnnotation(
        row_barplot = anno_oncoprint_barplot(c("AMP", "HOMDEL"),  # only AMP and HOMDEL
            border = TRUE, height = unit(4, "cm"), 
            axis_param = list(side = "bottom", labels_rot = 90))
    ),
    remove_empty_columns = TRUE, remove_empty_rows = TRUE,
    column_title = column_title, heatmap_legend_param = heatmap_legend_param,
    column_title_gp = gpar(fontsize = 9, fontface = "bold"))


ht1 + ht1 
```














what are the membrane genes

```{r}
head(downloadedData$subcellular_location)

table(downloadedData$subcellular_location$main_location) %>% 
  sort()

memberane_genes<- downloadedData$subcellular_location %>% 
  filter(str_detect(main_location, "Plasma membrane")) 

inner_join(data, memberane_genes, by = c("ensembl" = "ensembl")) 
```


Make a heatmap 

```{r}
library(ComplexHeatmap)

data2<- data %>% 
  filter(gene %in% memberane_genes$gene)

mat<- as.matrix(data2[, -c(1,2,3)])
rownames(mat)<- data2$gene

ComplexHeatmap::Heatmap(mat,
                        show_row_names = FALSE,
                        show_column_names = FALSE)
```